### 2016-03-08 paper/paper.tex
some things

\documentclass{article}
\usepackage{minted}
\usemintedstyle{emacs}

\begin{document}

% question: Can macros be used to create interesting and useful parallelism
% tools for recursive functions in clojure.
\section{Introduction}
% most important section in paper:
% - introduce the problem being solved
% - prove that this problem is interesting
% - why is this hard?
% - how does it differ from other approaches?
% - what are key results/limitations
Clojure is a lisp--like language running on the Java Virtual Machine with good support for concurrent and parallel programming.
Because Clojure is a lisp, it also has a powerful macro system.
Using Clojure's macro system, we have implemented a variety of macros which transform recursive serial code into equivalent parallel code, executing with the Java platform's ForkJoinPool executor service.

These macros are designed to allow developers to trivially utilize additional processing power, without doing much additional work.
The primary target audience is not developers concerned with extracting every last ounce of performance from their code, but instead developers trying to write highly expressive and readable code who would benefit from a speedup, but are not willing to rewrite substantial portions of their code, or write their code in a different manner to gain this speedup.
This means that these macros must be simple to use and must place very few of their decisions in the hands of the developers.
It also means that these macros must protect against invalid usage.

Unlike other approaches for simple parallelism in clojure, these macros parallelize control flow. % what's the opposite of data parallel
Many of the other Clojure parallelism constructs are data parallel, meaning that the parallelize operations on large sets of data. % example, pmap

The macros discussed in this paper only apply to a subset of the Clojure language. For example, code which manipulates state would be ill---suited for automatic parallelization with these macros.

This paper is divided into $n$ sections. The first discusses the macros we've implemented, along with the appropriate use cases for each. In the second section, we present benchmarks demonstrating the cases in which the parallelism macros can (and cannot) improve performance of existing code.

\section{Macros}
\subsection{parlet}
The first of the parallel macros is called parlet.
This macro is used to implement many of the others, so we will discuss it first.
Parlet is a modified version of Clojure let form.
Clojure's let form is used to bind names to values, then executes some other
Clojure forms in the new environment.
The parlet macro is exactly the same as a Clojure's let, but it evaluates all of
it's bindings in parallel (if there is no dependencies between them).
For example, suppose I had some long running function foo, if I needed to add
the result of two calls to this function, I could do the following:

\begin{minted}{clojure}
    (parlet
      [a (foo value1)
       b (foo value2)]
      (+ a b))
\end{minted}

This would execute the call to foo with value1 and the call to foo with value2
in parallel. The execution of each of these tasks is done in a ForkJoinPool.
Both tasks are submitted to the pool and told to execute in the background.
Then, in each form given to the let where the name is expected to have a value
bound to it, the name is replaced with a call to join for each new task. The
example above expands to:

\begin{minted}{clojure}
(let
 [a (fork (new-task (fn [] (foo v1))))
  b (fork (new-task (fn [] (foo v2))))]
 (+ (join a) (join b)))
\end{minted}

The joins are intentionally pushed into the forms in the let. Each join will
block until the computation finishes (if it has not already), so we push these
as far as we possibly can. This also means that nested parlets will benefit from
as much parallelism as possible. For example:

\begin{minted}{clojure}
    (parlet
      [a (foo 100)]
      ;; some other code not using a
      (parlet
        [b (foo 200)]
        (+ a b)))
\end{minted}

Expands to:

\begin{minted}{clojure}
(let
 [a (fork (new-task (fn [] (foo 100))))]
 ;; some other code not using a
 (let
  [b (fork (new-task (fn* ([] (foo 200)))))]
  (+ (join a) (join b))))
\end{minted}

Code like this may not arise when the code is human generated, but it may arise
when the code is generated by another macro. We will see some examples of this
later.

The parlet macro also supports dependency detection. Clojure let forms can use
names defined previously in the let, and the bindings are evaluated from first
to last. The following form is valid:

\begin{minted}{clojure}
    (let
     [a 1
      b (+ 1 a)]
     a)
\end{minted}

This form would evaluate to 2. If we plan on evaluating each binding in
parallel, we can't allow the bindings to have dependencies. So, the parlet macro
looks through the bindings in the macro to determine if any of them depend on
any of the names defined previously in the macro. If there are any, the macro
will halt the compiler and report an error to the user.

\subsection{parexpr}

Parexpr breaks down expressions and (aggressively) evaluates them in parallel.
parexpr uses parlet to parallelize forms. Suppose again that I had a long
running function foo, I wanted to call it twice, and add the results. With
parexpr the following will make both calls to foo in parallel:

\begin{minted}{clojure}
    (parexp (+ (foo value1) (foo value2)))
\end{minted}

The parlet macro expands to something like the following:

\begin{minted}{clojure}
(parlet
 [expr16512 (foo 11)
  expr16511 (foo 10)]
 (parlet
  [expr16513 (+ expr16511 expr16512)]
  expr16513))
\end{minted}

\end{document}

### 2016-03-08 run_remote.sh

@@ -1,0 +1,47 @@
+#!/bin/bash
+
+# wrapper around run suite that runs the suite remotely
+
+display_usage() {
+    echo -e "usage $0 host auto_parallel_local num_cpus local_out specs"
+    echo -e "where"
+    echo -e "\thost - remote host"
+    echo -e "\tauto_parallel_local - directory with auto parallel code locally"
+    echo -e "\tnum_cpus - number of cpus to use remotely"
+    echo -e "\tlocal_out - local directory to put output in"
+    echo -e "\tspecs - specs to run, relative to the auto_parallel root on the remote"
+}
+
+if [ $# -lt 4 ]
+then
+    display_usage
+    exit 1
+fi
+
+host=$1
+auto_parallel_dir_local=$2
+auto_parallel_dir_remote=/root/auto_parallel
+num_cpus=$3
+local_output_dir=$4
+
+# idk
+tmp=( "$@" )
+specs=( "${tmp[@]:3}" )
+specs=$(echo "${specs[*]}")
+
+echo $specs
+
+echo "pushing directory" $auto_parallel_dir_local
+
+# copy a dir to run something
+rsync -a $auto_parallel_dir_local root@$host:$(dirname $auto_parallel_dir_remote)
+
+ssh root@$host auto_parallel_dir_remote=$auto_parallel_dir_remote num_cpus=$num_cpus specs=\"$specs\" 'bash -s' <<'ENDSSH'
+    cd $auto_parallel_dir_remote
+    ./run_suite.sh $num_cpus out $specs
+ENDSSH
+
+echo "pulling directory" $auto_parallel_dir_remote/out
+
+# copy an output dir back
+rsync -a root@$host:$auto_parallel_dir_remote/out .

### 2016-03-17 run_suite.sh
removed cores logic from run suite and benchmark code

@@ -1,17 +1,13 @@
 #!/bin/bash
 
-# this script ignores hyperthreading and NUMA
-# if you need to care about those things, you'll have to be more careful
-
 display_usage() {
-    echo -e "usage $0 number_cpus output_dir specs_to_run"
+    echo -e "usage $0 output_dir specs_to_run"
     echo -e "where"
-    echo -e "\tnumber_cpus - number of cpus to allow each process to use"
     echo -e "\toutput_dir - directory to dump results to"
     echo -e "\tspecs_to_run - the remaining arguments (at least one) are benchmark specs to execute"
 }
 
-if [ $# -lt 3 ]
+if [ $# -lt 2 ]
 then
     display_usage
     exit 1
@@ -23,34 +19,27 @@ set -a
 source ./env
 set +a
 
-num_cpus=$1
-cpus=$(seq -s"," 0 $(expr $num_cpus - 1))
-
-output_dir=$2
-output_dir=$output_dir/$num_cpus-$(date +"%s")
+output_dir=$1
+output_dir=$output_dir/$(date +"%s")
 log_file=$output_dir/log
 
 tmp=( "$@" )
-specs=( "${tmp[@]:2}" )
+specs=( "${tmp[@]:1}" )
 
-echo "Running with $num_cpus cpus for each benchmark, and writing into $output_dir"
+echo "Running -- writing into $output_dir"
 
 # make a new directory for this run
 mkdir -p $output_dir
 cp env $output_dir/env
 
-echo "run starting" > $log_file
-date +"%m-%d-%y %H:%m:%S" >> $log_file
-date +"%s" >> $log_file
-echo "num_cpus: $num_cpus" >> $log_file
+echo "run starting" | tee $log_file
+date +"%m-%d-%y %H:%m:%S" | tee -a $log_file
+date +"%s" | tee -a $log_file
 
-echo >> $log_file
-echo "using cpus: $cpus" >> $log_file
-
-echo >> $log_file
-echo "running these specs:" >> $log_file
+echo | tee -a $log_file
+echo "running these specs:" | tee -a $log_file
 for spec in "${specs[@]}"; do
-    echo "$spec" >> $log_file
+    echo "$spec" | tee -a $log_file
 done
 
 lein compile | tee -a $log_file
@@ -62,7 +51,7 @@ for spec in "${specs[@]}"; do
     base=$output_dir/$base
 
     # run the spec and do the output
[…]

### 2016-03-17 build_machine.sh
try using vagrant to benchmark

@@ -1,0 +1,22 @@
+#!/bin/bash
+
+# get java
+wget --no-check-certificate --no-cookies --header "Cookie: oraclelicense=accept-securebackup-cookie" http://download.oracle.com/otn-pub/java/jdk/8u74-b02/jdk-8u74-linux-x64.tar.gz -nv
+
+tar xzf jdk-8u74-linux-x64.tar.gz
+rm jdk-8u74-linux-x64.tar.gz
+
+echo "export JAVA_HOME=~/jdk1.8.0_74" >> /home/vagrant/.bashrc
+echo "export PATH=\$JAVA_HOME/bin:\$PATH" >> /home/vagrant/.bashrc
+
+# get lein
+mkdir bin
+cd bin
+wget https://raw.githubusercontent.com/technomancy/leiningen/stable/bin/lein -nv
+chmod +x lein
+echo "export PATH=~/bin:\$PATH" >> /home/vagrant/.bashrc
+
+# get the code
+cd /home/vagrant
+apt-get install git -y
+git clone https://github.com/dpzmick/auto_parallel.git

### 2016-03-19 paper/paper.tex
added a breif blob about defparfun, improved into

@@
 % - what are key results/limitations
-Clojure is a lisp--like language running on the Java Virtual Machine with good support for concurrent and parallel programming.
-Because Clojure is a lisp, it also has a powerful macro system.
-Using Clojure's macro system, we have implemented a variety of macros which transform recursive serial code into equivalent parallel code, executing with the Java platform's ForkJoinPool executor service.
+Clojure is a lisp--like language running on the Java Virtual Machine.
+The language's default immutable data structures and it's sophisticated STM system make it well suited for parallel and concurrent programming.
+Additionally, because Clojure runs on the Java Virtual Machine, Clojure developers can take advantage of existing cross--platform parallelism libraries, such as Java's excellent ExecutorService framework, to write parallel code.
 
-These macros are designed to allow developers to trivially utilize additional processing power, without doing much additional work.
-The primary target audience is not developers concerned with extracting every last ounce of performance from their code, but instead developers trying to write highly expressive and readable code who would benefit from a speedup, but are not willing to rewrite substantial portions of their code, or write their code in a different manner to gain this speedup.
-This means that these macros must be simple to use and must place very few of their decisions in the hands of the developers.
-It also means that these macros must protect against invalid usage.
+Taking advantage of Clojure's parallel potential is not entirely straightforward, however.
+Interfacing with Java libraries directly is simple, but the interface code often feels slightly out of place when surrounded by other Clojure code.
+The STM system is easy to use, but the STM constructs are often too low level to be immediately useful.
+% for example, writing a parallel map with atoms would be a big pain
 
-Unlike other approaches for simple parallelism in clojure, these macros parallelize control flow. % what's the opposite of data parallel
-Many of the other Clojure parallelism constructs are data parallel, meaning that the parallelize operations on large sets of data. % example, pmap
+% is control flow the right word here? I feel like something else might be more appropriate.
+% Recursion is control flow, parlet is control flow, so I guess it makes sense.
+As a result, there are a variety of libraries designed to allow developers to take advantage of the parallelism and concurrency potential in the language.
+Many of these library functions and builtins are data-parallel; they are designed to apply some sort of operation to a set of data.
+This is often the kind of parallelism that developers desire, or are accustomed to identifying, but Clojure's nature as a functional language with immutable structures also makes it possible to exploit parallelism in the control flow of the code.
 
-The macros discussed in this paper only apply to a subset of the Clojure language. For example, code which manipulates state would be ill---suited for automatic parallelization with these macros.
+% even if it is expressed in control flow, it is still data parallelism probably
+Using Clojure's macro system, we have implemented a few macros which allow developers to take advantage of Clojure's parallelism potential when their parallelism is more directly expressed in the control flow of their application.
+We have shown that it is possible to attain noticeable degrees of parallelism with minimal code changes (with respect to serial code) through these macros, without the need for sophisticated dependency analysis (find OpenMP paper) often required in other parallelism systems.
+
+% the other existing libraries are also a testament to this statement
+% we sort of side stepped this by saying that the macros only work on pure code.
+% Need to demonstrate that idiotmatic clojure code is actually pure and that these macros actually make sense.
 
@@
 The first of the parallel macros is called parlet.
-This macro is used to implement many of the others, so we will discuss it first.
-Parlet is a modified version of Clojure let form.
-Clojure's let form is used to bind names to values, then executes some other
-Clojure forms in the new environment.
-The parlet macro is exactly the same as a Clojure's let, but it evaluates all of
+Parlet is a modified version of the Clojure let form.
+The parlet macro has exactly the same semitics as Clojure's let, but it evaluates all of
 it's bindings in parallel (if there is no dependencies between them).
 For example, suppose I had some long running function foo, if I needed to add
-the result of two calls to this function, I could do the following:
+the result of two calls to this function, I could do the following to execute my long running calls in parallel:
 
+% TODO use a figure
 \begin{minted}{clojure}
@@
        b (foo value2)]
+       ... ; some other code here
       (+ a b))
@@
 
-This would execute the call to foo with value1 and the call to foo with value2
-in parallel. The execution of each of these tasks is done in a ForkJoinPool.
-Both tasks are submitted to the pool and told to execute in the background.
-Then, in each form given to the let where the name is expected to have a value
-bound to it, the name is replaced with a call to join for each new task. The
-example above expands to:
+The execution of each of the function calls done in a ForkJoinPool.
+The calls to foo both start in the background, then we attempt to evaluate the body of the let.
+This means that the code in the body of the let which does not depend on the computations of a and b can execute without delay.
+This is obviously only safe when foo does not have side effects which may impact the other code in the function.
+In Clojure, it is very common (although not guaranteed by the compiler) for function calls to be pure.
 
-\begin{minted}{clojure}
-(let
- [a (fork (new-task (fn [] (foo v1))))
-  b (fork (new-task (fn [] (foo v2))))]
- (+ (join a) (join b)))
-\end{minted}
-
-The joins are intentionally pushed into the forms in the let. Each join will
-block until the computation finishes (if it has not already), so we push these
-as far as we possibly can. This also means that nested parlets will benefit from
-as much parallelism as possible. For example:
+Nested parlet macros will also behave quite well.
+For example:
 
@@
 
-Expands to:
+Both calls to the function foo will be processed in the background, until we reach the line using the variables that the results are bound to.
+This means that parlets can be nested with little concern (again as long as functions are pure).
 
-\begin{minted}{clojure}
-(let
- [a (fork (new-task (fn [] (foo 100))))]
- ;; some other code not using a
- (let
-  [b (fork (new-task (fn* ([] (foo 200)))))]
-  (+ (join a) (join b))))
-\end{minted}
+Code like this may not arise when the code is human generated, but it may arise when the code is generated by another macro.
+We will see some examples of this later.
 
-Code like this may not arise when the code is human generated, but it may arise
-when the code is generated by another macro. We will see some examples of this
-later.
-
-The parlet macro also supports dependency detection. Clojure let forms can use
-names defined previously in the let, and the bindings are evaluated from first
-to last. The following form is valid:
+The parlet macro also supports dependency detection.
+Clojure let forms can use names defined previously in the let, and the bindings are evaluated from first to last.
+The following form is valid (the binding to b used the value defined in the first binding):
 
@@
 
-This form would evaluate to 2. If we plan on evaluating each binding in
-parallel, we can't allow the bindings to have dependencies. So, the parlet macro
-looks through the bindings in the macro to determine if any of them depend on
-any of the names defined previously in the macro. If there are any, the macro
-will halt the compiler and report an error to the user.
+This form would evaluate to 2.
+If we plan on evaluating each binding in parallel, we can't allow the bindings to have dependencies.
+So, the parlet macro looks through the bindings in the macro to determine if any of them depend on
+any of the names defined previously in the macro.
+If there are any, the macro will halt the compiler and report an error to the user.
+
+This simple dependency check, as well as the commonplace purity in Clojure code, allow us to ensure correct parallelism with this simple macro.
+Because the Clojure compiler does not enforce purity, we still need to trust the programmer to validate the purity assumption before inserting the macro.
+% but lots of clojure code is already pure blah blah blah
 
@@
 Parexpr breaks down expressions and (aggressively) evaluates them in parallel.
-parexpr uses parlet to parallelize forms. Suppose again that I had a long
-running function foo, I wanted to call it twice, and add the results. With
-parexpr the following will make both calls to foo in parallel:
+parexpr uses parlet to parallelize forms.
+Suppose again that I had a long running function foo, I wanted to call it twice, and add the results.
+With parexpr the following will make both calls to foo in parallel:
 
@@
 
-The parlet macro expands to something like the following:
+The macro crawls the expression and expands every subexpression into a parlet.
+This macro has limited applicability, but, we can also convince ourselves of its correctness easily.
 
+\subsection{defparfun}
+
+The defparun macro is the most interesting of the 3 macros.
+defparfun allows a programmer to parallelize calls to a recursive function.
+This is best explained with an example:
+
+% TODO check my own stupid syntax
 \begin{minted}{clojure}
-(parlet
- [expr16512 (foo 11)
-  expr16511 (foo 10)]
- (parlet
-  [expr16513 (+ expr16511 expr16512)]
-  expr16513))
+    (defparfun fib [n] (> n 30)
+      (if (or (= n 1) (= n 2))
+          1
+          (+
+            (fib (- n 1))
+            (fib (- n 2)))))
 \end{minted}
 
+This defines a parallel Fibonacci function, which will only execute in parallel when the value of it's argument is greater than 30.
+
+% TODO actual explain this
+The macro first identifies any recursive call sites, then moves the function call as far ``up'' in the function as it can, without breaking any dependencies.
+Again, this dependency check is relatively simple.
+After this transformation, it is possible to replace the let forms which bind the function results to their values with parlet forms providing the same bindings.
+The introduction of the parlet form introduces parallelism, so each recursive call will execute in the ForkJoin pool, in parallel.
+Each of these subsequent recursive calls may create more tasks, which will also be executed in the pool.
+ForkJoin pools are designed to handle this type of computation, so the computation will proceed without blocking (as long as function is pure and performs no thread--blocking I/O).
+% TODO reference a ForkJoin pool paper
+
 \end{document}

### 2016-03-21 build_machine.sh
added vagrant stuff for experiment

@@ -1,22 +1,26 @@
 #!/bin/bash
 
+# roughly what needs to happen to set up a new machine
+sudo apt-get install git -y
+# sudo apt-get install vim -y
+
 # get java
 wget --no-check-certificate --no-cookies --header "Cookie: oraclelicense=accept-securebackup-cookie" http://download.oracle.com/otn-pub/java/jdk/8u74-b02/jdk-8u74-linux-x64.tar.gz -nv
 
 tar xzf jdk-8u74-linux-x64.tar.gz
 rm jdk-8u74-linux-x64.tar.gz
 
-echo "export JAVA_HOME=~/jdk1.8.0_74" >> /home/vagrant/.bashrc
-echo "export PATH=\$JAVA_HOME/bin:\$PATH" >> /home/vagrant/.bashrc
+echo "export JAVA_HOME=~/jdk1.8.0_74" > ~/.bashrc
+echo "export PATH=\$JAVA_HOME/bin:\$PATH" >> ~/.bashrc
 
 # get lein
 mkdir bin
 cd bin
 wget https://raw.githubusercontent.com/technomancy/leiningen/stable/bin/lein -nv
 chmod +x lein
-echo "export PATH=~/bin:\$PATH" >> /home/vagrant/.bashrc
+echo "export PATH=~/bin:\$PATH" >> ~/.bashrc
+echo "export LEIN_ROOT=1" >> ~/.bashrc
 
 # get the code
-cd /home/vagrant
-apt-get install git -y
+cd
 git clone https://github.com/dpzmick/auto_parallel.git

### 2016-03-21 run_remote.sh

@@ -1,16 +1,18 @@
 #!/bin/bash
 
-# wrapper around run suite that runs the suite remotely
-
 display_usage() {
-    echo -e "usage $0 host auto_parallel_local num_cpus local_out specs"
+    echo -e "usage $0 num_cpus num_boxes local_out specs"
     echo -e "where"
-    echo -e "\thost - remote host"
-    echo -e "\tauto_parallel_local - directory with auto parallel code locally"
-    echo -e "\tnum_cpus - number of cpus to use remotely"
+    echo -e "\tnum_cpus  - number of cpus in each vm"
+    echo -e "\tnum_boxes - number of vms to create"
     echo -e "\tlocal_out - local directory to put output in"
     echo -e "\tspecs - specs to run, relative to the auto_parallel root on the remote"
 }
+
+# configs
+ssh_config_file=/tmp/vagrant-ssh-cfg
+auto_parallel_dir_local=.
+auto_parallel_dir_remote=/home/vagrant/auto_parallel
 
 if [ $# -lt 4 ]
 then
@@ -18,30 +20,47 @@ then
     exit 1
 fi
 
-host=$1
-auto_parallel_dir_local=$2
-auto_parallel_dir_remote=/root/auto_parallel
-num_cpus=$3
-local_output_dir=$4
+num_cpus=$1
+num_boxes=$2
+local_out=$3
 
 # idk
 tmp=( "$@" )
 specs=( "${tmp[@]:3}" )
 specs=$(echo "${specs[*]}")
 
-echo $specs
+# bring up all the vms
+# save the hosts to the config file one at a time
+for n in `seq 1 $num_boxes` ; do
+    host="cores"$num_cpus"n"$n
+    vagrant up $host
+    vagrant ssh-config $host >> $ssh_config_file
+done
 
-echo "pushing directory" $auto_parallel_dir_local
+# run the benchmarks,
+for n in `seq 1 $num_boxes` ; do
+    host="cores"$num_cpus"n"$n
+    ( echo cd $auto_parallel_dir_remote ; echo git pull ; rm -rf out ; \
+        echo ./run_suite.sh out $specs) | \
+        { ssh -F $ssh_config_file $host 2>&1 | sed "s/^/$host==>/" ; } &
+done
+wait
 
-# copy a dir to run something
-rsync -a $auto_parallel_dir_local root@$host:$(dirname $auto_parallel_dir_remote)
[…]

### 2016-03-21 run_suite.sh

@@ -1,5 +1,8 @@
 #!/bin/bash
 
+# TODO log cpu usage
+# TODO log git commit hash
+
 display_usage() {
     echo -e "usage $0 output_dir specs_to_run"
     echo -e "where"
@@ -20,7 +23,7 @@ source ./env
 set +a
 
 output_dir=$1
-output_dir=$output_dir/$(date +"%s")
+output_dir=$output_dir/$(date +"%s")-$(hostname)
 log_file=$output_dir/log
 
 tmp=( "$@" )

### 2016-03-22 paper/paper.tex
please be here
fiddling with paper, using google to run everything

@@
-\documentclass{article}
+\documentclass{sig-alternate}
 \usepackage{minted}
 \usemintedstyle{emacs}
+
+\def\mintcode#1{\mbox{\mintinline{clojure}{#1}}}
+\def\code#1{\mbox{\texttt{#1}}}
+\def\parlet{\code{parlet}}
 
@@
 The language's default immutable data structures and it's sophisticated STM system make it well suited for parallel and concurrent programming.
-Additionally, because Clojure runs on the Java Virtual Machine, Clojure developers can take advantage of existing cross--platform parallelism libraries, such as Java's excellent ExecutorService framework, to write parallel code.
+Additionally, because Clojure runs on the Java Virtual Machine, Clojure developers can take advantage of existing cross--platform parallelism libraries, such as Java's excellent \code{ExecutorService} framework, to write parallel code.
 
@@
 \subsection{parlet}
-The first of the parallel macros is called parlet.
-Parlet is a modified version of the Clojure let form.
-The parlet macro has exactly the same semitics as Clojure's let, but it evaluates all of
-it's bindings in parallel (if there is no dependencies between them).
-For example, suppose I had some long running function foo, if I needed to add
-the result of two calls to this function, I could do the following to execute my long running calls in parallel:
+The first of the parallel macros is called \parlet{}.
+\parlet{} is a modified version of the Clojure let form.
+The \parlet{} macro has exactly the same semantics as Clojure's \code{let}, but it evaluates all of it's bindings in parallel.
+For example, suppose I had some long running function \code{foo}.
+I need to add the result of two calls to this function.
+In Figure \ref{fig:parlet_basic}, we use \parlet{} to make two calls to \code{foo}, then add the results.
 
-% TODO use a figure
+\begin{figure}[h]
     \begin{minted}{clojure}
@@
     \end{minted}
+    \caption{Example of a parlet form}
+    \label{fig:parlet_basic}
+\end{figure}
 
-The execution of each of the function calls done in a ForkJoinPool.
-The calls to foo both start in the background, then we attempt to evaluate the body of the let.
-This means that the code in the body of the let which does not depend on the computations of a and b can execute without delay.
-This is obviously only safe when foo does not have side effects which may impact the other code in the function.
+In this example, the expressions \mintcode{(foo value1)} and \mintcode{(foo value2)} are both evaluated as \code{ForkJoinTask}s in a \code{ForkJoinPool}.
+
+The calls to \code{foo} are both \code{forked} immediately, then we attempt to evaluate the body of the \code{let}.
+This means that the code in the body of the \code{let} which does not depend on the computations of \code{a} and \code{b} can execute without delay.
+
+This transformation is obviously only safe when \code{foo} does not have side effects which may impact the other code in the function, or the values of terms used in subsequent let terms.
 In Clojure, it is very common (although not guaranteed by the compiler) for function calls to be pure.
@@
 Nested parlet macros will also behave quite well.
-For example:
 
+\begin{figure}[h]
     \begin{minted}{clojure}
@@
     \end{minted}
+    \caption{Nested parlet forms}
+    \label{fig:parlet_noblock}
+\end{figure}
 
-Both calls to the function foo will be processed in the background, until we reach the line using the variables that the results are bound to.
+In Figure \ref{fig:parlet_noblock}, calls to the function foo will be processed in the background until we reach the line using the variables that the results are bound to.
 This means that parlets can be nested with little concern (again as long as functions are pure).
@@
 Clojure let forms can use names defined previously in the let, and the bindings are evaluated from first to last.
-The following form is valid (the binding to b used the value defined in the first binding):
 
+\begin{figure}[h]
     \begin{minted}{clojure}
-    (let
+        (parlet
         [a 1
@@
     \end{minted}
+    \caption{A parlet containing a dependency}
+    \label{fig:parlet_dep}
+\end{figure}
 
-This form would evaluate to 2.
+Without the parlet, the let form in Figure \ref{fig:parlet_dep} would evaluate to 2.
 If we plan on evaluating each binding in parallel, we can't allow the bindings to have dependencies.
@@
 Suppose again that I had a long running function foo, I wanted to call it twice, and add the results.
-With parexpr the following will make both calls to foo in parallel:
+The code in Figure \ref{fig:parexpr} will make both of the long running calls to foo in parallel.
 
+\begin{figure}[h]
     \begin{minted}{clojure}
@@
     \end{minted}
+    \caption{Example using parexpr to evaluate long running functions}
+    \label{fig:parexpr}
+\end{figure}
 
@@
 
+\section{Benchmarking}
+\subsection{Setup}
+
+\section{Results}
+
+\section{Conclusions}
+% the jvm is (mostly) awesome
+% immutability makes things like this really straightforward
+
 \end{document}

### 2016-03-22 run_remote.sh

@@ -10,9 +10,11 @@ display_usage() {
 }
 
 # configs
-ssh_config_file=/tmp/vagrant-ssh-cfg
+ssh_config_file=~/.ssh/google_compute_engine
 auto_parallel_dir_local=.
-auto_parallel_dir_remote=/home/vagrant/auto_parallel
+auto_parallel_dir_remote=/home/dpzmick/auto_parallel
+zone=us-central1-a
+project=senior-thesis-1257
 
 if [ $# -lt 4 ]
 then
@@ -23,6 +25,7 @@ fi
 num_cpus=$1
 num_boxes=$2
 local_out=$3
+mem=6
 
 # idk
 tmp=( "$@" )
@@ -30,37 +33,45 @@ specs=( "${tmp[@]:3}" )
 specs=$(echo "${specs[*]}")
 
 # bring up all the vms
-# save the hosts to the config file one at a time
 for n in `seq 1 $num_boxes` ; do
     host="cores"$num_cpus"n"$n
-    vagrant up $host
-    vagrant ssh-config $host >> $ssh_config_file
+
+    gcloud compute instances create $host --custom-cpu $num_cpus \
+        --custom-memory $mem --image ap-base-image &
+
+    # vagrant up $host
+    # vagrant ssh-config $host >> $ssh_config_file
 done
+wait
+
+# get the ssh things configured
+gcloud compute config-ssh
 
 # run the benchmarks,
 for n in `seq 1 $num_boxes` ; do
     host="cores"$num_cpus"n"$n
-    ( echo cd $auto_parallel_dir_remote ; echo git pull ; rm -rf out ; \
+    ( echo cd $auto_parallel_dir_remote ; echo git pull ; echo rm -rf out ; \
         echo ./run_suite.sh out $specs) | \
-        { ssh -F $ssh_config_file $host 2>&1 | sed "s/^/$host==>/" ; } &
+        { gcloud compute ssh $host 2>&1 | sed "s/^/$host==>/" ; } &
 done
 wait
 
 # copy the results into the local directory
 for n in `seq 1 $num_boxes` ; do
     host="cores"$num_cpus"n"$n
-    mkdir -p $local_out/$host
-    scp -F $ssh_config_file -r $host:$auto_parallel_dir_remote/out $local_out/$host
+
+    my_out=$local_out/$host-$(date +"%s")
+    mkdir -p $my_out
+    gcloud compute instances list --format=json > $my_out/gcloud.json
[…]

### 2016-03-22 vis/dir2csv.py

@@ -3,7 +3,10 @@ from __future__ import print_function
 import sys
 from sys import stderr
 from os import listdir
+import json
+import re
 import os.path
+import csv
 
 # converts a directory of output files into a csv file with timings for each run
 # output csv should be in form:
@@ -74,8 +77,8 @@ def handle_trial(input_dir, trial_name):
         print("skipping", file=stderr)
         return []
 
-    cpus_line = filter(lambda l: l.startswith('num_cpus'), lines)
-    num_cpus = int(cpus_line[0].split(":")[1])
+    # cpus_line = filter(lambda l: l.startswith('num_cpus'), lines)
+    # num_cpus = int(cpus_line[0].split(":")[1])
 
     date = int(lines[2])
 
@@ -88,8 +91,57 @@ def handle_trial(input_dir, trial_name):
 
     my_data = []
     for name, time in children_data:
-        my_data.append( (name, num_cpus, date, time) )
+        my_data.append({
+            'spec-name': name,
+            'date': date,
+            'mean-runtime': time
+            })
+
     return my_data
+
+def handle_host(host_dir):
+    print("host: ", host_dir, file=stderr)
+
+    try:
+
+        # get the number of cores this host has
+        mejson = None
+        with open(os.path.join(host_dir, 'gcloud.json')) as me:
+            mejson = json.loads(me.read())
+
+        mename = None
+        with open(os.path.join(host_dir, 'me')) as me:
+            mename = me.readlines()[0].strip()
+
+        me = None
+        for d in mejson:
+            if d[u'name'] == mename:
+                me = d
+                break
+
+        plaform = me['cpuPlatform']
+        # might as well get the number of cores from the name, since its string
+        # manip to get it from the json anyway
+        # of course, find the biggest hack possible and use that
+        cores = int(re.sub(r"\D", "", mename.split('n')[0]))
+
+        # for this host, get all the data from the benchmark directory
+        bdir = os.path.join(host_dir, 'out')
+
+        trials = []
+        for path in listdir(bdir):
+            trials.extend(handle_trial(bdir, path))
+
+        for t in trials:
+            t['platform'] = plaform
+            t['host']     = mename
+            t['cores']    = cores
+
+        return trials
+
+    except Exception, e:
+        print(str(e), file=stderr)
+        return []
 
 if __name__ == "__main__":
[…]

### 2016-03-23 paper/paper.tex
updated intro

@@
 \documentclass{sig-alternate}
+% \usepackage[none]{hyphenat}
+\usepackage{hyperref}
+
 \usepackage{minted}
 \usemintedstyle{emacs}
+
+\bibliographystyle{plain}
 
@@
 
+\author{
+    Zmick, David\\
+    \texttt{zmick2@illinois.edu}
+}
+
+\title{Macros for straightforward parallelism in Clojure}
+
 \begin{document}
+
+\maketitle
 
@@
 \section{Introduction}
-% most important section in paper:
-% - introduce the problem being solved
-% - prove that this problem is interesting
-% - why is this hard?
-% - how does it differ from other approaches?
-% - what are key results/limitations
 Clojure is a lisp--like language running on the Java Virtual Machine.
-The language's default immutable data structures and it's sophisticated STM system make it well suited for parallel and concurrent programming.
-Additionally, because Clojure runs on the Java Virtual Machine, Clojure developers can take advantage of existing cross--platform parallelism libraries, such as Java's excellent \code{ExecutorService} framework, to write parallel code.
+The language's default immutable data structures and it's sophisticated Software Transactional Memory (STM) system make it well suited for parallel programming.
+Because Clojure runs on the Java Virtual Machine, Clojure developers can take advantage of existing cross--platform parallelism libraries, such as Java's excellent \code{ExecutorService} framework, to write parallel code.
 
+% for example, writing a parallel map with atoms would be a big pain
 Taking advantage of Clojure's parallel potential is not entirely straightforward, however.
-Interfacing with Java libraries directly is simple, but the interface code often feels slightly out of place when surrounded by other Clojure code.
-The STM system is easy to use, but the STM constructs are often too low level to be immediately useful.
-% for example, writing a parallel map with atoms would be a big pain
+STM has proven to be very successful as a clear construct for concurrent programming\cite{Jones2007a}, but these constructs are often too low level to be of much use to developers whose central concern is not parallelism\cite{Boehm2009}.
 
-% is control flow the right word here? I feel like something else might be more appropriate.
-% Recursion is control flow, parlet is control flow, so I guess it makes sense.
-As a result, there are a variety of libraries designed to allow developers to take advantage of the parallelism and concurrency potential in the language.
+As a result, there are a variety of libraries designed to allow developers to take advantage of the parallelism potential in Clojure.\footnotemark{}\footnotemark{}\footnotemark{}\footnotemark{}\footnotemark{}
 Many of these library functions and builtins are data-parallel; they are designed to apply some sort of operation to a set of data.
-This is often the kind of parallelism that developers desire, or are accustomed to identifying, but Clojure's nature as a functional language with immutable structures also makes it possible to exploit parallelism in the control flow of the code.
+This is often the kind of parallelism that developers desire\cite{Okur2012a} but Clojure's nature as a functional language with immutable structures also makes it possible to easily exploit control parallelism (also known as task parallelism\cite{Andradea}).
+
+\footnotetext[1]{\url{https://github.com/clojure/core.async}}
+\footnotetext[2]{\url{https://github.com/TheClimateCorporation/claypoole}}
+\footnotetext[3]{\url{https://clojuredocs.org/clojure.core/pmap}}
+\footnotetext[4]{\url{http://clojure.org/reference/reducers}}
+\footnotetext[5]{\url{https://github.com/aphyr/tesser}}
 
 % even if it is expressed in control flow, it is still data parallelism probably
-Using Clojure's macro system, we have implemented a few macros which allow developers to take advantage of Clojure's parallelism potential when their parallelism is more directly expressed in the control flow of their application.
-We have shown that it is possible to attain noticeable degrees of parallelism with minimal code changes (with respect to serial code) through these macros, without the need for sophisticated dependency analysis (find OpenMP paper) often required in other parallelism systems.
+Using Clojure's macro system, we have implemented a set of macros which allow developers to take advantage of Clojure's parallelism potential when their existing code is written such that parallelism is exposed through control flow.
+We have shown that it is possible to attain noticeable degrees of parallelism with minimal code changes (with respect to serial code), without the need for sophisticated dependency analysis  often required in other parallelism systems.
 
@@
 % Need to demonstrate that idiotmatic clojure code is actually pure and that these macros actually make sense.
-
-This paper is divided into $n$ sections.
-The first discusses the macros we've implemented, along with the appropriate use cases for each.
-In the second section, we present benchmarks demonstrating the cases in which the parallelism macros can (and cannot) improve performance of existing code.
 
@@
 
-In this example, the expressions \mintcode{(foo value1)} and \mintcode{(foo value2)} are both evaluated as \code{ForkJoinTask}s in a \code{ForkJoinPool}.
+In this example, the expressions \mintcode{(foo value1)} and \mintcode{(foo value2)} are both evaluated as \code{ForkJoinTask}s in a \code{ForkJoinPool}\cite{Lea2000}.
 
@@
 
+\bibliography{/home/dpzmick/Documents/Bibtex/senior-thesis.bib}
+
 \end{document}

### 2016-03-23 vis/dir2csv.py

@@ -66,38 +66,41 @@ def handle_trial(input_dir, trial_name):
     with open(os.path.join(path, 'log')) as log_f:
         lines = log_f.readlines()
 
-    if lines == None:
-        print("didn't get any lines in the log for", path, file=stderr)
-        return []
+    try:
 
-    lines = map(lambda l: l.strip(), lines)
+        if lines == None:
+            print("didn't get any lines in the log for", path, file=stderr)
+            return []
 
-    if "finished test" not in lines:
-        print("didn't finish the test:", path, file=stderr)
-        print("skipping", file=stderr)
-        return []
+        lines = map(lambda l: l.strip(), lines)
 
-    # cpus_line = filter(lambda l: l.startswith('num_cpus'), lines)
-    # num_cpus = int(cpus_line[0].split(":")[1])
+        if "finished test" not in lines:
+            print("didn't finish the test:", path, file=stderr)
+            print("skipping", file=stderr)
+            return []
 
-    date = int(lines[2])
+        date = int(lines[2])
 
-    children_data = []
-    for f in listdir(path):
-        if f == 'log' or f == 'env':
-            continue
+        children_data = []
+        for f in listdir(path):
+            if f == 'log' or f == 'env':
+                continue
 
-        children_data.append(handle_single_spec(path, f))
+            children_data.append(handle_single_spec(path, f))
 
-    my_data = []
-    for name, time in children_data:
-        my_data.append({
-            'spec-name': name,
-            'date': date,
-            'mean-runtime': time
-            })
+        my_data = []
+        for name, time in children_data:
+            my_data.append({
+                'spec-name': name,
+                'date': date,
+                'mean-runtime': time
+                })
 
-    return my_data
+        return my_data
+
+    except Exception, e:
+        print(str(e), file=stderr)
+        return {}
 
 def handle_host(host_dir):
     print("host: ", host_dir, file=stderr)

### 2016-03-23 run_suite.sh
update to runsuite

@@ -47,6 +47,8 @@ done
 
 lein compile | tee -a $log_file
 
+cat /proc/cpuinfo | tee -a $log_file
+
 # each spec gets its own output file
 for spec in "${specs[@]}"; do
     # setup filenames

### 2016-03-24 vis/basic_plot.R
more data crunching stuff

@@ -1,0 +1,72 @@
+require(data.table)
+require(plyr)
+require(Hmisc)
+
+# gets all speedups for a single host
+# host-id, cores, speedup, spec
+host_data_for_benchmark <- function(host, benchmark, dat, parfun_type) {
+  host_data <- dat[dat$host.id == host & dat$spec.name %like% benchmark,]
+  
+  important <- c("cores", "host.id", "platform", "mean.runtime")
+  
+  serial_speeds   <- host_data[host_data$spec.name %like% "serial",important]
+  parallel_speeds <- host_data[host_data$spec.name %like% parfun_type,important]
+  
+  # I already know these are for the same type of benchmark, since I filtered already
+  d <- merge(serial_speeds, parallel_speeds, by=c("cores", "host.id", "platform"))
+  
+  # x is the serial one, y is the parallel speed
+  d$speedup <- d$mean.runtime.x / d$mean.runtime.y
+  
+  d$benchmark <- rep(benchmark, nrow(d))
+  
+  return(d)
+}
+
+host_speedups_for_bench <- function(data, benchmark, parfun_type) {
+  # for every host that we find, compute their average runtime on each test
+  hosts <- unique(data$host.id)
+  
+  curr <- host_data_for_benchmark(hosts[1], benchmark, data, parfun_type)
+  for (host in tail(hosts, -1)) {
+    d <- host_data_for_benchmark(host, benchmark, data, parfun_type)
+    curr <- rbind(curr, d)
+  }
+  
+  return(curr)
+}
+
+plot_means_error_for_benchmark <- function(data, benchmark, parfun_type) {
+  d <- host_speedups_for_bench(data, benchmark, parfun_type)[,c("cores", "speedup")]
+  means <- ddply(d, "cores", colwise(mean))
+  sds <- ddply(d, "cores", colwise(sd))
+  sumr <- merge(means, sds, "cores", suffixes=c(".mean", ".sd"))
+  
+  max_mean = max(unlist(lapply(sumr$speedup.mean, function(x) if(is.na(x)) -Inf else x)))
+  max_sd   = max(unlist(lapply(sumr$speedup.sd+sumr$speedup.mean, function(x) if(is.na(x)) -Inf else x)))
+  
+  plot(sumr$cores, sumr$speedup.mean, ylim=c(0,max(max_sd, max_mean)), xlab = "", ylab = "")
+  
+  title(
[…]

### 2016-03-24 vis/dir2csv.py

@@ -123,6 +123,8 @@ def handle_host(host_dir):
                 break
 
         plaform = me['cpuPlatform']
+        meid    = me['id']
+
         # might as well get the number of cores from the name, since its string
         # manip to get it from the json anyway
         # of course, find the biggest hack possible and use that
@@ -139,6 +141,7 @@ def handle_host(host_dir):
             t['platform'] = plaform
             t['host']     = mename
             t['cores']    = cores
+            t['host-id']  = meid
 
         return trials
 

### 2016-03-24 paper/paper.tex
not sure what changes made, cleaning up local repo

@@
 Clojure is a lisp--like language running on the Java Virtual Machine.
-The language's default immutable data structures and it's sophisticated Software Transactional Memory (STM) system make it well suited for parallel programming.
+The language's default immutable data structures and sophisticated Software Transactional Memory (STM) system make it well suited for parallel programming.
 Because Clojure runs on the Java Virtual Machine, Clojure developers can take advantage of existing cross--platform parallelism libraries, such as Java's excellent \code{ExecutorService} framework, to write parallel code.
@@
 Taking advantage of Clojure's parallel potential is not entirely straightforward, however.
-STM has proven to be very successful as a clear construct for concurrent programming\cite{Jones2007a}, but these constructs are often too low level to be of much use to developers whose central concern is not parallelism\cite{Boehm2009}.
+STM has proven to be very successful as a clear construct for concurrent programming\cite{Jones2007a}, but these constructs are often too low level to be of much use to developers whose central concerns are not parallelism\cite{Boehm2009}.
 
@@
 Many of these library functions and builtins are data-parallel; they are designed to apply some sort of operation to a set of data.
-This is often the kind of parallelism that developers desire\cite{Okur2012a} but Clojure's nature as a functional language with immutable structures also makes it possible to easily exploit control parallelism (also known as task parallelism\cite{Andradea}).
+Developers have a good relationship with data parallel problems\cite{Okur2012a}, but Clojure's nature as a functional language with immutable structures also makes it possible to easily exploit control parallelism (also known as task parallelism\cite{Andradea, Rodr}).
 
@@
 We have shown that it is possible to attain noticeable degrees of parallelism with minimal code changes (with respect to serial code), without the need for sophisticated dependency analysis  often required in other parallelism systems.
+% indicate that we cooperate with STM and other clojure constructs
 

### 2016-03-25 vis/basic_plot.R
updated these things

@@ -38,24 +38,16 @@ host_speedups_for_bench <- function(data, benchmark, parfun_type) {
 
 plot_means_error_for_benchmark <- function(data, benchmark, parfun_type) {
   d <- host_speedups_for_bench(data, benchmark, parfun_type)[,c("cores", "speedup")]
+  
+  boxplot(speedup ~ cores, data=d,
+          main=paste(parfun_type, "speedup for", benchmark, "from", nrow(d), "trials", sep=" "),
+          xlab="number of cores",
+          ylab="speedup w.r.t serial version"
+  )
+  
   means <- ddply(d, "cores", colwise(mean))
   sds <- ddply(d, "cores", colwise(sd))
   sumr <- merge(means, sds, "cores", suffixes=c(".mean", ".sd"))
-  
-  max_mean = max(unlist(lapply(sumr$speedup.mean, function(x) if(is.na(x)) -Inf else x)))
-  max_sd   = max(unlist(lapply(sumr$speedup.sd+sumr$speedup.mean, function(x) if(is.na(x)) -Inf else x)))
-  
-  plot(sumr$cores, sumr$speedup.mean, ylim=c(0,max(max_sd, max_mean)), xlab = "", ylab = "")
-  
-  title(
-    paste(parfun_type, "speedup for", benchmark, sep=" "),
-    xlab="number of cores",
-    ylab="speedup w.r.t serial version")
-  
-  with (
-    data = sumr
-    , expr = errbar(cores, speedup.mean, speedup.mean+speedup.sd, speedup.mean-speedup.sd, add=T, pch=1, cap=.1)
-  )
   
   return(sumr)
 }

### 2016-03-25 vis/dir2csv.py

@@ -57,6 +57,8 @@ def handle_single_spec(trial_dir, spec_name):
 
 
 def handle_trial(input_dir, trial_name):
+    print("trial: ", trial_name, file=stderr)
+
     path = os.path.join(input_dir, trial_name)
     trial_date = trial_name.split('-')[1]
 
@@ -105,49 +107,47 @@ def handle_trial(input_dir, trial_name):
 def handle_host(host_dir):
     print("host: ", host_dir, file=stderr)
 
-    try:
+    # get the number of cores this host has
+    mejson = None
+    with open(os.path.join(host_dir, 'gcloud.json')) as me:
+        mejson = json.loads(me.read())
 
-        # get the number of cores this host has
-        mejson = None
-        with open(os.path.join(host_dir, 'gcloud.json')) as me:
-            mejson = json.loads(me.read())
+    mename = None
+    with open(os.path.join(host_dir, 'me')) as me:
+        mename = me.readlines()[0].strip()
 
-        mename = None
-        with open(os.path.join(host_dir, 'me')) as me:
-            mename = me.readlines()[0].strip()
+    me = None
+    for d in mejson:
+        if d[u'name'] == mename:
+            me = d
+            break
 
-        me = None
-        for d in mejson:
-            if d[u'name'] == mename:
-                me = d
-                break
+    plaform = me['cpuPlatform']
+    meid    = me['id']
 
-        plaform = me['cpuPlatform']
-        meid    = me['id']
+    # might as well get the number of cores from the name, since its string
+    # manip to get it from the json anyway
+    # of course, find the biggest hack possible and use that
+    cores = int(re.sub(r"\D", "", mename.split('n')[0]))
 
-        # might as well get the number of cores from the name, since its string
-        # manip to get it from the json anyway
-        # of course, find the biggest hack possible and use that
-        cores = int(re.sub(r"\D", "", mename.split('n')[0]))
+    # for this host, get all the data from the benchmark directory
+    bdir = os.path.join(host_dir, 'out')
 
-        # for this host, get all the data from the benchmark directory
-        bdir = os.path.join(host_dir, 'out')
+    trials = []
+    for path in listdir(bdir):
+        trials.extend(handle_trial(bdir, path))
 
-        trials = []
-        for path in listdir(bdir):
-            trials.extend(handle_trial(bdir, path))
+    for t in trials:
+        t['platform'] = plaform
+        t['host']     = mename
[…]

### 2016-04-02 paper/paper.tex
added java fib code + cleaning up

@@
 Using Clojure's macro system, we have implemented a set of macros which allow developers to take advantage of Clojure's parallelism potential when their existing code is written such that parallelism is exposed through control flow.
-We have shown that it is possible to attain noticeable degrees of parallelism with minimal code changes (with respect to serial code), without the need for sophisticated dependency analysis  often required in other parallelism systems.
+We have shown that it is possible to attain reasonable degrees of parallelism with minimal code changes (with respect to serial code).
+These transformation can be applied to idiomatic Clojure code without the need for sophisticated dependency analysis often required in other parallelism systems.
 % indicate that we cooperate with STM and other clojure constructs

### 2016-04-02 scripts/run_java_suite.sh

@@ -1,0 +1,48 @@
+#!/bin/bash
+
+# TODO log cpu usage
+# TODO log git commit hash
+
+display_usage() {
+    echo -e "usage $0 output_dir"
+    echo -e "where"
+    echo -e "\toutput_dir - directory to dump results to"
+}
+
+if [ $# -lt 1 ]
+then
+    display_usage
+    exit 1
+fi
+
+# first, load the environment needed for benchmarks
+# the set -a thing is to force them to export
+set -a
+source ./env
+set +a
+
+output_dir=$1
+output_dir=$output_dir/$(date +"%s")-$(hostname)
+log_file=$output_dir/log
+
+echo "Running -- writing into $output_dir"
+
+# make a new directory for this run
+mkdir -p $output_dir
+cp env $output_dir/env
+touch $output_dir/is_java
+
+echo "run starting" | tee $log_file
+date +"%m-%d-%y %H:%m:%S" | tee -a $log_file
+date +"%s" | tee -a $log_file
+
+echo | tee -a $log_file
+
+lein javac | tee -a $log_file
+
+cat /proc/cpuinfo | tee -a $log_file
+
+# run the spec and do the output
+lein jbenchmark | tee $log_file.bench
+
+echo "finished test" >> $log_file

### 2016-04-02 scripts/run_remote.sh

@@ -1,5 +1,7 @@
 #!/bin/bash
 
+# should never be run from the scripts directory
+
 display_usage() {
     echo -e "usage $0 num_cpus num_boxes local_out specs"
     echo -e "where"

### 2016-04-02 scripts/run_while_away.sh

@@ -1,0 +1,18 @@
+#!/bin/bash
+
+./run_remote.sh 1 23 out-tmp benchmark_specs/fib/big/{serial.csv,parfun.csv}
+./run_remote.sh 2 12 out-tmp benchmark_specs/fib/big/{serial.csv,parfun.csv}
+./run_remote.sh 4 6 out-tmp benchmark_specs/fib/big/{serial.csv,parfun.csv}
+./run_remote.sh 6 4 out-tmp benchmark_specs/fib/big/{serial.csv,parfun.csv}
+./run_remote.sh 1 23 out-tmp benchmark_specs/fib/big/{serial.csv,parfun.csv}
+./run_remote.sh 2 12 out-tmp benchmark_specs/fib/big/{serial.csv,parfun.csv}
+./run_remote.sh 4 6 out-tmp benchmark_specs/fib/big/{serial.csv,parfun.csv}
+./run_remote.sh 6 4 out-tmp benchmark_specs/fib/big/{serial.csv,parfun.csv}
+./run_remote.sh 1 23 out-tmp benchmark_specs/fib/big/{serial.csv,parfun.csv}
+./run_remote.sh 2 12 out-tmp benchmark_specs/fib/big/{serial.csv,parfun.csv}
+./run_remote.sh 4 6 out-tmp benchmark_specs/fib/big/{serial.csv,parfun.csv}
+./run_remote.sh 6 4 out-tmp benchmark_specs/fib/big/{serial.csv,parfun.csv}
+./run_remote.sh 1 23 out-tmp benchmark_specs/fib/big/{serial.csv,parfun.csv}
+./run_remote.sh 2 12 out-tmp benchmark_specs/fib/big/{serial.csv,parfun.csv}
+./run_remote.sh 4 6 out-tmp benchmark_specs/fib/big/{serial.csv,parfun.csv}
+./run_remote.sh 6 4 out-tmp benchmark_specs/fib/big/{serial.csv,parfun.csv}

### 2016-04-02 src_java/com/dpzmick/auto_parallel_java/App.java

@@ -1,0 +1,72 @@
+package com.dpzmick.auto_parallel_java;
+
+import java.util.concurrent.RecursiveTask;
+import java.util.concurrent.ForkJoinPool;
+
+import com.google.caliper.Runner;
+import com.google.caliper.SimpleBenchmark;
+
+class FibTask extends RecursiveTask<Integer> {
+    private int grain_;
+    private int n_;
+
+    public FibTask(int n, int grain) {
+        grain_ = grain;
+        n_     = n;
+    }
+
+    static Integer serialCompute(int n) {
+        if (n == 0 || n == 1)
+            return 1;
+
+        return serialCompute(n - 1) + serialCompute(n - 2);
+    }
+
+    @Override
+    public Integer compute() {
+        if (n_ == 0 || n_ == 1) {
+            return 1;
+        }
+
+        if (n_ <= grain_) {
+            return serialCompute(n_);
+        }
+
+        FibTask f1 = new FibTask(n_ - 1, grain_);
+        FibTask f2 = new FibTask(n_ - 2, grain_);
+        f1.fork();
+        return f1.join() + f2.compute();
+    }
+}
+
+
+public class App {
+    public static class Benchmark extends SimpleBenchmark {
+        private int n;
+
+        @Override
+        protected void setUp() throws Exception {
+            n = Integer.parseInt(System.getenv("BIG_FIB"));
+        }
+
+        public void timeFibFJ(int reps) {
+            for (int i = 0; i < reps; i++) {
+                ForkJoinPool pool = new ForkJoinPool();
+                FibTask f = new FibTask(n, 31);
+                Integer res = pool.invoke(f);
+                System.out.println(res);
+            }
+        }
+
+        public void timeFibSerial(int reps) {
+            for (int i = 0; i < reps; i++) {
+                System.out.println(FibTask.serialCompute(n));
+            }
+        }
+
+    }
+
+    public static void main(String[] args) {
+        Runner.main(Benchmark.class, new String[0]);
+    }
+}

### 2016-04-02 scripts/run_java_remote.sh
keep the args in the data, run java remotely

@@ -3,12 +3,11 @@
 # should never be run from the scripts directory
 
 display_usage() {
-    echo -e "usage $0 num_cpus num_boxes local_out specs"
+    echo -e "usage $0 num_cpus num_boxes local_out"
     echo -e "where"
     echo -e "\tnum_cpus  - number of cpus in each vm"
     echo -e "\tnum_boxes - number of vms to create"
     echo -e "\tlocal_out - local directory to put output in"
-    echo -e "\tspecs - specs to run, relative to the auto_parallel root on the remote"
 }
 
 # configs
@@ -18,7 +17,7 @@ auto_parallel_dir_remote=/home/dpzmick/auto_parallel
 zone=us-central1-a
 project=senior-thesis-1257
 
-if [ $# -lt 4 ]
+if [ $# -lt 3 ]
 then
     display_usage
     exit 1
@@ -28,11 +27,6 @@ num_cpus=$1
 num_boxes=$2
 local_out=$3
 mem=6
-
-# idk
-tmp=( "$@" )
-specs=( "${tmp[@]:3}" )
-specs=$(echo "${specs[*]}")
 
 # bring up all the vms
 for n in `seq 1 $num_boxes` ; do
@@ -53,7 +47,7 @@ gcloud compute config-ssh
 for n in `seq 1 $num_boxes` ; do
     host="cores"$num_cpus"n"$n
     ( echo cd $auto_parallel_dir_remote ; echo git pull ; echo rm -rf out ; \
-        echo ./run_suite.sh out $specs) | \
+        echo scripts/run_java_suite.sh out) | \
         { gcloud compute ssh $host 2>&1 | sed "s/^/$host==>/" ; } &
 done
 wait

### 2016-04-02 vis/dir2csv.py

@@ -53,7 +53,11 @@ def handle_single_spec(trial_dir, spec_name):
 
     mean_time = to_micros(mean_time_base, mean_time_unit)
 
-    return (spec_name, mean_time)
+    # get the args
+    args_line = filter(lambda l: l.startswith("running"), lines)[0]
+    args = re.findall('\((.*)\)', args_line)[0]
+
+    return (spec_name, mean_time, args.split(' '))
 
 
 def handle_trial(input_dir, trial_name):
@@ -91,12 +95,17 @@ def handle_trial(input_dir, trial_name):
             children_data.append(handle_single_spec(path, f))
 
         my_data = []
-        for name, time in children_data:
-            my_data.append({
+        for name, time, args in children_data:
+            medat = {
                 'spec-name': name,
                 'date': date,
                 'mean-runtime': time
-                })
+                }
+
+            for i, arg in enumerate(args):
+                medat['args' + str(i)] = arg
+
+            my_data.append(medat)
 
         return my_data
 

### 2016-04-03 scripts/run_remote.sh
gonna try a bigger fib

@@ -53,7 +53,7 @@ gcloud compute config-ssh
 for n in `seq 1 $num_boxes` ; do
     host="cores"$num_cpus"n"$n
     ( echo cd $auto_parallel_dir_remote ; echo git pull ; echo rm -rf out ; \
-        echo ./run_suite.sh out $specs) | \
+        echo scripts/run_suite.sh out $specs) | \
         { gcloud compute ssh $host 2>&1 | sed "s/^/$host==>/" ; } &
 done
 wait

### 2016-04-03 src_java/com/dpzmick/auto_parallel_java/App.java
fixed java

@@ -35,7 +35,11 @@ class FibTask extends RecursiveTask<Integer> {
         FibTask f1 = new FibTask(n_ - 1, grain_);
         FibTask f2 = new FibTask(n_ - 2, grain_);
         f1.fork();
-        return f1.join() + f2.compute();
+
+        Integer this_thread = f2.compute();
+        Integer other_thread = f1.join();
+
+        return this_thread + other_thread;
     }
 }
 
@@ -67,6 +71,12 @@ public class App {
     }
 
     public static void main(String[] args) {
-        Runner.main(Benchmark.class, new String[0]);
+        // Runner.main(Benchmark.class, new String[0]);
+        System.out.println("hello world");
+        ForkJoinPool pool = new ForkJoinPool();
+        int n = Integer.parseInt(System.getenv("BIG_FIB"));
+        FibTask f = new FibTask(n, 31);
+        Integer res = pool.invoke(f);
+        System.out.println(res);
     }
 }

### 2016-04-03 src_java/com/dpzmick/auto_parallel_java/App.java
pls

@@ -71,12 +71,12 @@ public class App {
     }
 
     public static void main(String[] args) {
-        // Runner.main(Benchmark.class, new String[0]);
-        System.out.println("hello world");
-        ForkJoinPool pool = new ForkJoinPool();
-        int n = Integer.parseInt(System.getenv("BIG_FIB"));
-        FibTask f = new FibTask(n, 31);
-        Integer res = pool.invoke(f);
-        System.out.println(res);
+        Runner.main(Benchmark.class, new String[0]);
+        // System.out.println("hello world");
+        // ForkJoinPool pool = new ForkJoinPool();
+        // int n = Integer.parseInt(System.getenv("BIG_FIB"));
+        // FibTask f = new FibTask(n, 31);
+        // Integer res = pool.invoke(f);
+        // System.out.println(res);
     }
 }

### 2016-04-05 src_java/com/dpzmick/auto_parallel_java/App.java
waaaaaaaaaaaaaahhhhhhhhhhhh

@@ -56,7 +56,7 @@ public class App {
         public void timeFibFJ(int reps) {
             for (int i = 0; i < reps; i++) {
                 ForkJoinPool pool = new ForkJoinPool();
-                FibTask f = new FibTask(n, 31);
+                FibTask f = new FibTask(n, 35);
                 Integer res = pool.invoke(f);
                 System.out.println(res);
             }

### 2016-04-05 vis/dir2csv.py

@@ -57,7 +57,46 @@ def handle_single_spec(trial_dir, spec_name):
     args_line = filter(lambda l: l.startswith("running"), lines)[0]
     args = re.findall('\((.*)\)', args_line)[0]
 
+    variance_lines = filter(lambda l: "variance" in l.lower(), lines)
+    if len(variance_lines) != 0:
+        vline = variance_lines[0]
+        if "severely" in vline.lower():
+            print("ignored ", trial_dir, spec_name, file=stderr)
+            return None
+
     return (spec_name, mean_time, args.split(' '))
+
+def java_trial(input_dir, trial_name):
+    # trial is in log.bench
+    lines = []
+    with open(os.path.join(input_dir, trial_name, "log.bench")) as bench:
+        lines = bench.readlines()
+
+    unit_line = filter(lambda l: "runtime" in l.lower(), lines)[0]
+    unit = unit_line.split()[1]
+
+    # need to get the serial time and the FJ time
+    fj_line = filter(lambda l: "FibFJ" in l, lines)[1]
+    fj_time = to_micros(float(fj_line.split()[1]), unit)
+
+    serial_line = filter(lambda l: "FibSerial" in l, lines)[1]
+    serial_time = to_micros(float(serial_line.split()[1]), unit)
+
+    m1 = {
+        'spec-name': 'fib.fj_java',
+        'mean-runtime': fj_time,
+        'args0': None,
+        'date': None
+    }
+
+    m2 = {
+        'spec-name': 'fib.serial_java',
+        'mean-runtime': serial_time,
+        'args0': None,
+        'date': None
+    }
+
+    return [m1, m2]
 
 
 def handle_trial(input_dir, trial_name):
@@ -92,7 +131,11 @@ def handle_trial(input_dir, trial_name):
             if f == 'log' or f == 'env':
                 continue
 
-            children_data.append(handle_single_spec(path, f))
+            d = handle_single_spec(path, f)
+            if d == None:
+                return []
+
+            children_data.append(d)
 
         my_data = []
         for name, time, args in children_data:
@@ -144,7 +187,10 @@ def handle_host(host_dir):
 
     trials = []
     for path in listdir(bdir):
-        trials.extend(handle_trial(bdir, path))
+        if 'is_java' in listdir(os.path.join(bdir, path)):
+            trials.extend(java_trial(bdir, path))
+        else:
+            trials.extend(handle_trial(bdir, path))
 
     for t in trials:
[…]

### 2016-04-05 paper/paper.Rnw
knitr

--- 2016-04-02 paper/paper.tex
@@
 \section{Introduction}
+Some text
+<<RTest>>=
+2+2+2
+@
+
 Clojure is a lisp--like language running on the Java Virtual Machine.
@@
 % for example, writing a parallel map with atoms would be a big pain
-Taking advantage of Clojure's parallel potential is not entirely straightforward, however.
+However, taking advantage of Clojure's parallel potential is not entirely straightforward.
 STM has proven to be very successful as a clear construct for concurrent programming\cite{Jones2007a}, but these constructs are often too low level to be of much use to developers whose central concerns are not parallelism\cite{Boehm2009}.

### 2016-04-05 paper/go.sh

@@ -1,0 +1,1 @@
+latexmk paper.Rnw -shell-escape -pdf -pvc

### 2016-04-05 paper/paper.Rnw
got knitr working

@@
-\documentclass{sig-alternate}
-% \usepackage[none]{hyphenat}
+\documentclass{article}
 \usepackage{hyperref}
@@
 \begin{document}
+<<include=FALSE>>=
+source("../vis/basic_plot.R")
+data <- read.csv("../newopts.csv")
+@
 
@@
 \section{Introduction}
-Some text
-<<RTest>>=
-2+2+2
-@
 
@@
 \section{Results}
+<<javafib, fig.pos="H", fig.cap="Fibonacci Java Performance", echo=F>>=
+jfib <- plot_means_error_for_benchmark(data, "fib", "fj")
+@
+
+<<fibparfun, fig.pos="H", fig.cap="Fibonacci Clojure (defparfun) Performance", echo=F>>=
+cljfib <- plot_means_error_for_benchmark(data, "fib", "parfun")
+@
 

### 2016-04-05 paper/go.sh

@@ -1,1 +1,3 @@
-latexmk paper.Rnw -shell-escape -pdf -pvc
+set -e
+Rscript -e "require(knitr); knit('paper.Rnw')"
+latexmk paper.tex -shell-escape -pdf -pv

### 2016-04-05 vis/basic_plot.R

@@ -40,7 +40,7 @@ plot_means_error_for_benchmark <- function(data, benchmark, parfun_type) {
   d <- host_speedups_for_bench(data, benchmark, parfun_type)[,c("cores", "speedup")]
   
   boxplot(speedup ~ cores, data=d,
-          main=paste(parfun_type, "speedup for", benchmark, "from", nrow(d), "trials", sep=" "),
+          main=paste("speedup from", nrow(d), "trials", sep=" "),
           xlab="number of cores",
           ylab="speedup w.r.t serial version"
   )
@@ -60,5 +60,5 @@ plot_runtimes_scatter <- function(data, benchmark, parfun_type) {
   points(serial, col="red")
 }
 
-data <- read.csv("data.csv", header=TRUE, sep=",")
+# data <- read.csv("data.csv", header=TRUE, sep=",")
 # plot_means_error_for_benchmark(dat, "id3.small", "parfun")

### 2016-04-05 vis/dir2csv.py
fixed dir2csv

@@ -34,7 +34,11 @@ def to_micros(b, unit):
     raise Exception("fixmepls")
 
 def handle_single_spec(trial_dir, spec_name):
-    print(trial_dir, spec_name, file=stderr)
+    print("single_spec: ", trial_dir, spec_name, file=stderr)
+
+    if "make_data" in spec_name:
+        print("returning none", file=stderr)
+        return None
 
     lines = None
     with open(os.path.join(trial_dir, spec_name)) as f:
@@ -133,7 +137,7 @@ def handle_trial(input_dir, trial_name):
 
             d = handle_single_spec(path, f)
             if d == None:
-                return []
+                continue
 
             children_data.append(d)
 

### 2016-04-08 paper/paper.Rnw
updated paper

@@
 \usepackage{hyperref}
+\usepackage{tikz}
+\usetikzlibrary{shapes.geometric, arrows}
+\usetikzlibrary{positioning}
 
@@
 \def\parlet{\code{parlet}}
+\def\parexpr{\code{parexpr}}
+\def\defparfun{\code{defparfun}}
 
@@
 source("../vis/basic_plot.R")
-data <- read.csv("../newopts.csv")
+data <- read.csv("../data.csv")
 @
@@
 
-% even if it is expressed in control flow, it is still data parallelism probably
-Using Clojure's macro system, we have implemented a set of macros which allow developers to take advantage of Clojure's parallelism potential when their existing code is written such that parallelism is exposed through control flow.
-We have shown that it is possible to attain reasonable degrees of parallelism with minimal code changes (with respect to serial code).
-These transformation can be applied to idiomatic Clojure code without the need for sophisticated dependency analysis often required in other parallelism systems.
-% indicate that we cooperate with STM and other clojure constructs
+Using Clojure's macro system, I have implemented a set of macros which allow developers to take advantage of Clojure's parallelism potential when their existing code is written such that parallelism is exposed through control flow.
+I have shown that it is possible to attain reasonable degrees of parallelism with minimal code changes, with respect to serial code, using these macros.
+These transformation can be applied to idiomatic Clojure code without the need for sophisticated dependency analysis often required in other parallelism systems, and do not inhibit the programmer's ability to leverage other Clojure concurrency constructs (such as the STM system).
+
+The intended users of these macros are developers whose primary concern is not performance, but may benefit from a simple mechanism with which they can take advantage of the many cores in their machines.
+Developers who are extremely concerned with performance and want a high degree of control should turn elsewhere (perhaps even to Java) to write their highly tuned code.
 
@@
 
-\section{Macros}
-\subsection{parlet}
+\section{Mostly Pure Functions}
+Before discussing the macros I've implemented, I need to loosely define a ``mostly pure'' function.
+A mostly pure function is a function that has no side effects which directly impact the values of other user-defined values, at the current level of abstraction.
+Mostly pure functions can be reordered without impacting the values of user variables, although a change in order may impact I/O behavior and output order of a program.
+When a programmer feels that it may be acceptable to change the order of mostly pure functions, we can reorder them subject to these constraints:
+
+\begin{enumerate}
+    \item A call to a mostly pure function $f$ in a block $B$ in a function's control flow graph can safely be moved to a block $P$ for which all paths in the graph though $P$ also go through $B$.
+    \item All of the arguments to the function are available at any block $P$ which is a candidate location for the function call.
+\end{enumerate}
+
+The first constraint is introduced so that we avoid making network requests or print statements which never would have originally occurred.
+We are only allowing these functions to be reordered within blocks where they would previously have been executed.
+Figures \ref{fig:move_mostly_pure} and \ref{fig:move_mostly_pure_through_let} demonstrate safe and unsafe positions a mostly pure function call can be moved to.
+
+\tikzstyle{goodnode} = [rectangle, draw=green, text width=8.8cm]
+\tikzstyle{badnode}  = [rectangle, draw=red,   text width=8.8cm]
+
+\begin{figure}[h]
+\begin{tikzpicture}[node distance=0.7cm,auto,>=stealth',->]
+    \node[coordinate] (0) at (1.0,0) {};
+    \node (start)       [badnode]                                              {\mintcode{(if (pred? a b c ...)}};
+    \node (iftruedo)    [badnode, below of=start, right of=start]              {\mintcode{(do}};
+    \node (iftruedo1)   [badnode, below of=iftruedo, right of=iftruedo]        {\mintcode{;; some code that uses b and/or c}};
+    \node (iftruedolet) [badnode, below of=iftruedo1,right of=iftruedo1]       {\mintcode{(let [a (bar b c)]}};
+    \node (iftruelet1)  [goodnode, below of=iftruedolet, right of=iftruedolet] {\mintcode{;; some code that uses a, b, and/or c}};
+    \node (iftruelet2)  [goodnode, below of=iftruelet1]                        {\mintcode{(foo a b c)))}};
+    \node (iffalse)     [badnode, below=4cm of 0]                              {\mintcode{(bar a b c))}};
+\end{tikzpicture}
+\caption{The call to the mostly pure function \code{foo} can only be moved to nodes which are {\color{green} green}}
+\label{fig:move_mostly_pure}
+\end{figure}
+
+\begin{figure}[h]
+\begin{tikzpicture}[node distance=0.7cm,auto,>=stealth',->]
+    \node[coordinate] (0) at (1.0,0) {};
+    \node (start)       [badnode]                                 {\mintcode{(do}};
+    \node (a)           [badnode, below of=start, right of=start] {\mintcode{(let [a (bar 10)]}};
+    \node (b)           [goodnode, below of=a, right of=a]        {\mintcode{;; some code that uses a}};
+    \node (c)           [goodnode, below of=b]                    {\mintcode{(let [b (bar a)]}};
+    \node (d)           [goodnode, below of=c, right of=c]        {\mintcode{;; some code that uses a and/or b}};
+    \node (e)           [goodnode, below of=d]                    {\mintcode{(foo a)))}};
+\end{tikzpicture}
+\caption{The call to the mostly pure function \code{foo} can only be moved to nodes which are {\color{green} green}}
+\label{fig:move_mostly_pure_through_let}
+\end{figure}
+
+Many functions fit this definition.
+For example, it usually does not matter when I call \code{println} as long as all of the of the arguments are available, and I don't call it in some totally unexpected location.
+This concept is not rigorously defined as there is not a good definition for the semantics of such functions.
+Instead, in this paper, we rely on the judgment of the programmer to tell us when a function should be considered mostly pure and when it should not be.
+Luckily, in Clojure, many functions are mostly pure, by convention.
+
+% A mostly pure function is similar to a \code{const} function in C++.
+% \code{const} functions promise not to modify any members of the object on which they are called (unless the member was declared \code{mutable}), essentially promising that the object will not change, from the user of the object's perspective, when the \code{const} function is called.
+
+\section{parlet}
 The first of the parallel macros is called \parlet{}.
-\parlet{} is a modified version of the Clojure let form.
-The \parlet{} macro has exactly the same semantics as Clojure's \code{let}, but it evaluates all of it's bindings in parallel.
+\parlet{} is a modified version of the Clojure \code{let} form.
+The \parlet{} macro has exactly the same semantics as Clojure's \code{let}, but it evaluates all of its bindings in parallel.
 For example, suppose I had some long running function \code{foo}.
@@
 In this example, the expressions \mintcode{(foo value1)} and \mintcode{(foo value2)} are both evaluated as \code{ForkJoinTask}s in a \code{ForkJoinPool}\cite{Lea2000}.
+Briefly, a \code{ForkJoinPool} is an executor for lightweight (often recursive) tasks.
+The execution engine uses a work stealing scheduler to schedule the lightweight tasks across a thread pool.
 
@@
 
-This transformation is obviously only safe when \code{foo} does not have side effects which may impact the other code in the function, or the values of terms used in subsequent let terms.
-In Clojure, it is very common (although not guaranteed by the compiler) for function calls to be pure.
-
-Nested parlet macros will also behave quite well.
+Since the \code{ForkJoinPool} is designed for recursive workloads, it allows tasks which are currently executing to create new tasks, submit them to the pool, then wait for the task to complete.
+None of the pool's threads will block when this occurs.
+This means that nested \parlet{} forms (Figure \ref{fig:parlet_noblock}) will execute without blocking.
 
@@
 
-In Figure \ref{fig:parlet_noblock}, calls to the function foo will be processed in the background until we reach the line using the variables that the results are bound to.
+In Figure \ref{fig:parlet_noblock}, calls to the function \code{foo} will be processed in the background until we reach the line using the variables that the results are bound to.
 This means that parlets can be nested with little concern (again as long as functions are pure).
@@
 
-The parlet macro also supports dependency detection.
-Clojure let forms can use names defined previously in the let, and the bindings are evaluated from first to last.
+\subsection{Dependencies}
+The \parlet{} macro also supports simple dependency detection.
+Clojure \code{let} forms can use names defined previously in the let, and the bindings are evaluated from first to last.
 
@@
 
-Without the parlet, the let form in Figure \ref{fig:parlet_dep} would evaluate to 2.
+Without the \parlet{}, the let form in Figure \ref{fig:parlet_dep} would evaluate to 2.
 If we plan on evaluating each binding in parallel, we can't allow the bindings to have dependencies.
-So, the parlet macro looks through the bindings in the macro to determine if any of them depend on
-any of the names defined previously in the macro.
+So, the \parlet{} macro looks through the bindings in the macro to determine if any of them depend on any of the names defined previously in the macro.
 If there are any, the macro will halt the compiler and report an error to the user.
 
-This simple dependency check, as well as the commonplace purity in Clojure code, allow us to ensure correct parallelism with this simple macro.
-Because the Clojure compiler does not enforce purity, we still need to trust the programmer to validate the purity assumption before inserting the macro.
-% but lots of clojure code is already pure blah blah blah
+This transformation is only safe when \code{foo} (and more generally, and function call in the bindings) is a mostly pure function.
+Here, we punt the definition of mostly pure to the programmer.
+If the programmer chooses to use a \parlet{} form, we assume that the functions are mostly pure.
+This simple dependency check, as well as the programmers promise that all function called are mostly pure Clojure code, allow us to ensure correct parallelism with this macro.
 
-\subsection{parexpr}
+\section{parexpr}
 
-Parexpr breaks down expressions and (aggressively) evaluates them in parallel.
-parexpr uses parlet to parallelize forms.
-Suppose again that I had a long running function foo, I wanted to call it twice, and add the results.
-The code in Figure \ref{fig:parexpr} will make both of the long running calls to foo in parallel.
+The \parexpr{} macro breaks down expressions and (aggressively) evaluates them in parallel.
+Suppose again that I had a long running function \code{foo} which I wanted to call twice, and add the results.
+The code in Figure \ref{fig:parexpr} will make both of the long running calls to \code{foo} in parallel.
+The \parexpr{} macro crawls the expression and expands it into multiple nested \parlet{} forms, filling each \parlet{} with as many evaluations as it can.
 
@@
 
-The macro crawls the expression and expands every subexpression into a parlet.
-This macro has limited applicability, but, we can also convince ourselves of its correctness easily.
+\section{defparfun}
 
-\subsection{defparfun}
-
-The defparun macro is the most interesting of the 3 macros.
-defparfun allows a programmer to parallelize calls to a recursive function.
+The \defparfun{} macro is the most interesting of the 3 macros.
+\defparfun{} allows a programmer to parallelize calls to a recursive function.
 This is best explained with an example:
@@
 % TODO check my own stupid syntax
+\begin{figure}[h]
     \begin{minted}{clojure}
@@
     \end{minted}
+    \caption{Example using defparfun to define a fibonacci function}
+    \label{fig:fib_parfun}
+\end{figure}
 
-This defines a parallel Fibonacci function, which will only execute in parallel when the value of it's argument is greater than 30.
+This defines a parallel Fibonacci function, which will only execute in parallel when the value of it's argument is greater than $30$.
 
-% TODO actual explain this
-The macro first identifies any recursive call sites, then moves the function call as far ``up'' in the function as it can, without breaking any dependencies.
-Again, this dependency check is relatively simple.
+\subsection{Imposed Constraints}
+The objective of the \defparfun{} macro is to enable simple, predictable parallelism for recursive functions which are mostly pure.
+We also aim to support flexible functions, but we cannot promise correctness when the macro is used with flexible functions.
+
+A mostly pure function call may be moved to a node $P$ in the control flow graph for the function iff every path through $P$ in the original function would execute the function call, and if the arguments to the function call are all available in $P$.
+We do not allow function calls to be introduced on a path which they would not have existed on before to preserve
+
+% we can move these functions around as long as all of their arguments are not changed
+% - the output order of the program may change slightly, but the results will not
+% don't move all calls to the top because don't want to create unpredictable performance or do wasteful evaluations, or break dependencies.
+
+\subsubsection{how it works or something}
+The macro first identifies any recursive call sites.
+It then introduces a name for the call, then pushes the evaluation of the function all as far ``up'' in the function as it can.
+We stop pushing a function call higher in a function when we encounter an \code{if} form, or we encounter any existing \code{let} expression which introduces a value that the function call depends on.
+We can think of this operation as an operation on a single ``basic block'' in the function.
+The function call can be freely moved inside of the function as long as the function call does not introduce any new behavior to the function.
+
 After this transformation, it is possible to replace the let forms which bind the function results to their values with parlet forms providing the same bindings.
@@
 ForkJoin pools are designed to handle this type of computation, so the computation will proceed without blocking (as long as function is pure and performs no thread--blocking I/O).
-% TODO reference a ForkJoin pool paper
 
 \section{Benchmarking}
-\subsection{Setup}
+To run benchmarks, we used Google's Cloud Compute product, because it allowed easy creation of machines with various configurations.
+Each benchmark was done on a virtual machine created to run the given benchmark.
+After the benchmark completed, the machines were destroyed.
+This allowed us to control the number of cores the JVM would use on every machine, since there is no good way to specify the number of cores the JVM will use for every thread it creates (including its garbage collection threads).
+A variety of small problems were implemented with serial recursive code, then invocations of the macros were added to the serial code.
+For any given test, the difference in the serial code and the parallel code is usually about 10 characters.
+I will present the results from a subset of the tests here.
+Each test was tested with a variety of different configurations and parallelism granularities where chosen to try and effectively use the resources available.
+For the tests with a fixed granularity (eg $n < 30$) the granularity was held constant for all tests.
+Some tests have dynamic granularity computed with respect to the number of cores available and the size of the input.
+For these tests, the granularity was not fixed.
 
-\section{Results}
+\subsection{Fibonacci}
+First we will look at the classical recursive fibonacci example.
+This benchmark was implemented in both Clojure and Java, to compare the speedups gotten with the macros vs the speedup gotten with handwritten Java Fork/Join code.
+Figure \ref{fig:javafib} displays the average speedup over many tests for the Java Fork/Join code.
+As we can see, we only get about a 3x speedup with 6 cores, in the simple Java implementation.
+
 <<javafib, fig.pos="H", fig.cap="Fibonacci Java Performance", echo=F>>=
@@
 cljfib <- plot_means_error_for_benchmark(data, "fib", "parfun")
+@
+
+<<id3parfun, fig.pos="H", fig.cap="id3 Clojure (defparfun) Performance", echo=F>>=
+id3 <- plot_means_error_for_benchmark(data, "id3", "parfun")
 @

### 2016-04-12 paper/paper.Rnw
updated paper

@@
-\documentclass{article}
+%\documentclass[12pt, letterpapper]{proc}
+\documentclass[12pt, letterpapper, titlepage]{article}
 \usepackage{hyperref}
@@
 \usepackage{minted}
-\usemintedstyle{emacs}
+%\usemintedstyle{emacs}
+\usemintedstyle{borland}
 
+\usepackage{subcaption}
+
+\usepackage[nottoc,numbib]{tocbibind}
 \bibliographystyle{plain}
@@
 source("../vis/basic_plot.R")
-data <- read.csv("../data.csv")
+data <- read.csv("../newnew.csv")
 @
@@
 \maketitle
+
+% \tableofcontents
+% \pagebreak
 
@@
 
-As a result, there are a variety of libraries designed to allow developers to take advantage of the parallelism potential in Clojure.\footnotemark{}\footnotemark{}\footnotemark{}\footnotemark{}\footnotemark{}
-Many of these library functions and builtins are data-parallel; they are designed to apply some sort of operation to a set of data.
+As a result, there are a variety of libraries designed to allow developers to take advantage of the parallelism potential in Clojure.
+Clojure builtins such as pmap\footnotemark[1] and the new reducers\footnotemark[2] library provide data parallel sequence manipulation and transformation functions.
+Third---party libraries like Tesser\footnotemark[3] and Claypoole\footnotemark[4] provide more data parallel APIs with slightly different goals than the builtin functions.
 Developers have a good relationship with data parallel problems\cite{Okur2012a}, but Clojure's nature as a functional language with immutable structures also makes it possible to easily exploit control parallelism (also known as task parallelism\cite{Andradea, Rodr}).
+An astute reader may observe that both Claypoole and the Clojure standard library include task parallel functions.
+These will be discussed in the Section \ref{label:related_work}
 
-\footnotetext[1]{\url{https://github.com/clojure/core.async}}
-\footnotetext[2]{\url{https://github.com/TheClimateCorporation/claypoole}}
-\footnotetext[3]{\url{https://clojuredocs.org/clojure.core/pmap}}
-\footnotetext[4]{\url{http://clojure.org/reference/reducers}}
-\footnotetext[5]{\url{https://github.com/aphyr/tesser}}
+\footnotetext[1]{\url{https://clojuredocs.org/clojure.core/pmap}}
+\footnotetext[2]{\url{http://clojure.org/reference/reducers}}
+\footnotetext[3]{\url{https://github.com/aphyr/tesser}}
+\footnotetext[4]{\url{https://github.com/TheClimateCorporation/claypoole}}
 
@@
 I have shown that it is possible to attain reasonable degrees of parallelism with minimal code changes, with respect to serial code, using these macros.
-These transformation can be applied to idiomatic Clojure code without the need for sophisticated dependency analysis often required in other parallelism systems, and do not inhibit the programmer's ability to leverage other Clojure concurrency constructs (such as the STM system).
-
 The intended users of these macros are developers whose primary concern is not performance, but may benefit from a simple mechanism with which they can take advantage of the many cores in their machines.
@@
 
-% the other existing libraries are also a testament to this statement
-% we sort of side stepped this by saying that the macros only work on pure code.
-% Need to demonstrate that idiotmatic clojure code is actually pure and that these macros actually make sense.
+\section{Related Work} \label{label:related_work}
+% only discuss task based parallelism in this section
+
+\subsection{Clojure and other lisps}
+% lparallel
+% futures
+% claypoole futures
+Makes lots of threads, unatural blah.
+Note sure what's bad about lparallel other than it isn't implemented in Clojure.
+Clojure also has access to Java's libraries, like ForkJoin.... but hard to use and not really a part of the Clojure language.
+
+Briefly, a \code{ForkJoinPool} is an executor for lightweight (often recursive) tasks.
+The execution engine uses a work stealing scheduler to schedule the lightweight tasks across a thread pool.
+
+\subsection{Other languages}
+% - cilk/tbb/openmp
+Compiler support is bad.
+Performance is quite good.
+Only real downside is that these are not implemented in Clojure.
+ForkJoin pools are essentially Cilk \cite{CilkReference, Frigo1998}.
+Both provide the support and runtime needed to easily parallelize code, but neither give programmers the ability to mark a function as ``parallel'' then let the macro system deal with granularity and the (small) syntactic overhead (cilk has very little extra syntax).
 
@@
 A mostly pure function is a function that has no side effects which directly impact the values of other user-defined values, at the current level of abstraction.
-Mostly pure functions can be reordered without impacting the values of user variables, although a change in order may impact I/O behavior and output order of a program.
-When a programmer feels that it may be acceptable to change the order of mostly pure functions, we can reorder them subject to these constraints:
+Mostly pure functions can be reordered (or interleaved) without impacting the values of user variables, although the change may impact I/O behavior and output order of a program.
+In some cases, the order of certain side effects may not matter to a programmer.
+For example, it may not matter if we reorder \mintcode{(download file1)} and \mintcode{(download file2)} for a programmer writing a web scraper, but it may matter to a programmer writing a I/O constrained server.
+When a programmer feels that it may be acceptable to change the order of execution of calls to mostly pure functions (the call is ``portable''), we can reorder them subject to these constraints:
 
@@
     \item A call to a mostly pure function $f$ in a block $B$ in a function's control flow graph can safely be moved to a block $P$ for which all paths in the graph though $P$ also go through $B$.
+        Figure \ref{fig:move_mostly_pure} provides and example of this constraint.
     \item All of the arguments to the function are available at any block $P$ which is a candidate location for the function call.
+        See Figure \ref{fig:move_mostly_pure_through_let} for an example.
 \end{enumerate}
 
-The first constraint is introduced so that we avoid making network requests or print statements which never would have originally occurred.
-We are only allowing these functions to be reordered within blocks where they would previously have been executed.
-Figures \ref{fig:move_mostly_pure} and \ref{fig:move_mostly_pure_through_let} demonstrate safe and unsafe positions a mostly pure function call can be moved to.
+The first constraint is introduced so that we avoid network requests or print statements which never would have originally occurred along any given path of execution.
+We do not want to allow reordering which introduces new computations or would result in unpredictable performance.
+The second constraint ensures that we don't ever violate the most basic of correctness properties.
+A more detailed algorithm for finding safe locations for portable mostly pure functions in Clojure code is discussed in Section \ref{label:mpure_algo}
 
-\tikzstyle{goodnode} = [rectangle, draw=green, text width=8.8cm]
-\tikzstyle{badnode}  = [rectangle, draw=red,   text width=8.8cm]
+\tikzstyle{goodnode} = [rectangle, draw=blue, text width=8.8cm]
+\tikzstyle{badnode}  = [rectangle, draw=none,  text width=8.8cm]
 
-\begin{figure}[h]
-\begin{tikzpicture}[node distance=0.7cm,auto,>=stealth',->]
-    \node[coordinate] (0) at (1.0,0) {};
-    \node (start)       [badnode]                                              {\mintcode{(if (pred? a b c ...)}};
+\begin{figure}[H]
+\centering
+\begin{tikzpicture}[node distance=0.6cm,auto]
+    \node[coordinate] (0) at (0.7,0) {};
+    \node (start0)      [badnode]                                        {\mintcode{;; we cannot move the call outside of the if.} \\ \mintcode{;; There is a path on which the call will never occur}};
+    \node (start)       [badnode, below of=start0]                       {\mintcode{(if (pred? a b c ...)}};
     \node (iftruedo)    [badnode, below of=start, right of=start]        {\mintcode{(do}};
-    \node (iftruedo1)   [badnode, below of=iftruedo, right of=iftruedo]        {\mintcode{;; some code that uses b and/or c}};
-    \node (iftruedolet) [badnode, below of=iftruedo1,right of=iftruedo1]       {\mintcode{(let [a (bar b c)]}};
-    \node (iftruelet1)  [goodnode, below of=iftruedolet, right of=iftruedolet] {\mintcode{;; some code that uses a, b, and/or c}};
-    \node (iftruelet2)  [goodnode, below of=iftruelet1]                        {\mintcode{(foo a b c)))}};
-    \node (iffalse)     [badnode, below=4cm of 0]                              {\mintcode{(bar a b c))}};
+    \node (iftruelet1)  [goodnode, below of=iftruedo, right of=iftruedo] {\mintcode{;; some code that uses a, b, and/or c}};
+    \node (iftruelet2)  [goodnode, below of=iftruelet1]                  {\mintcode{(foo a b c))}};
+    \node (iffalse)     [badnode, below=2.8cm of 0]                      {\mintcode{(bar a b c))}};
 \end{tikzpicture}
-\caption{The call to the mostly pure function \code{foo} can only be moved to nodes which are {\color{green} green}}
+\caption{The call to the mostly pure function \code{foo} can only be moved to the boxed nodes}
 \label{fig:move_mostly_pure}
@@
 
-\begin{figure}[h]
-\begin{tikzpicture}[node distance=0.7cm,auto,>=stealth',->]
+\begin{figure}[H]
+\centering
+\begin{tikzpicture}[node distance=0.6cm,auto]
     \node[coordinate] (0) at (1.0,0) {};
@@
 \end{tikzpicture}
-\caption{The call to the mostly pure function \code{foo} can only be moved to nodes which are {\color{green} green}}
+\caption{The call to the mostly pure function \code{foo} can only be moved to the boxed nodes}
 \label{fig:move_mostly_pure_through_let}
@@
 
-Many functions fit this definition.
-For example, it usually does not matter when I call \code{println} as long as all of the of the arguments are available, and I don't call it in some totally unexpected location.
-This concept is not rigorously defined as there is not a good definition for the semantics of such functions.
-Instead, in this paper, we rely on the judgment of the programmer to tell us when a function should be considered mostly pure and when it should not be.
-Luckily, in Clojure, many functions are mostly pure, by convention.
+Many Clojure functions fit this definition due to Clojure programming conventions and default immutable data structures.
+This concept is not rigorously defined as there is not a good definition for the semantics of such functions. % TODO find references
+Instead, in this paper, we rely on the judgment of the programmer to tell us when a mostly pure function is portable.
 
@@
 \begin{figure}[h]
+\centering
+\RecustomVerbatimEnvironment{Verbatim}{BVerbatim}{}
 \begin{minted}{clojure}
@@
 In this example, the expressions \mintcode{(foo value1)} and \mintcode{(foo value2)} are both evaluated as \code{ForkJoinTask}s in a \code{ForkJoinPool}\cite{Lea2000}.
-Briefly, a \code{ForkJoinPool} is an executor for lightweight (often recursive) tasks.
-The execution engine uses a work stealing scheduler to schedule the lightweight tasks across a thread pool.
 
@@
 \defparfun{} allows a programmer to parallelize calls to a recursive function.
-This is best explained with an example:
-
-% TODO check my own stupid syntax
-\begin{figure}[h]
-    \begin{minted}{clojure}
-        (defparfun fib [n] (> n 30)
-          (if (or (= n 1) (= n 2))
-              1
-              (+
-                (fib (- n 1))
-                (fib (- n 2)))))
-    \end{minted}
-    \caption{Example using defparfun to define a fibonacci function}
-    \label{fig:fib_parfun}
-\end{figure}
-
+For an example see Figure \ref{fig:fib_code}.
 This defines a parallel Fibonacci function, which will only execute in parallel when the value of it's argument is greater than $30$.
 
-\subsection{Imposed Constraints}
+\subsection{Imposed Constraints} \label{label:mpure_algo}
 The objective of the \defparfun{} macro is to enable simple, predictable parallelism for recursive functions which are mostly pure.
@@
 
-\subsubsection{how it works or something}
+\subsection{Moving portable mostly pure functions}
 The macro first identifies any recursive call sites.
@@
 
-After this transformation, it is possible to replace the let forms which bind the function results to their values with parlet forms providing the same bindings.
-The introduction of the parlet form introduces parallelism, so each recursive call will execute in the ForkJoin pool, in parallel.
+After this transformation, it is possible to replace the let forms which bind the function results to their values with \parlet{} forms providing the same bindings.
+The introduction of the parlet form introduces parallelism, so each recursive call will execute in the \code{ForkJoin} pool, in parallel.
 Each of these subsequent recursive calls may create more tasks, which will also be executed in the pool.
-ForkJoin pools are designed to handle this type of computation, so the computation will proceed without blocking (as long as function is pure and performs no thread--blocking I/O).
+\code{ForkJoin} pools are designed to handle this type of computation, so the computation will proceed without blocking (as long as function is pure and performs no thread--blocking I/O).
 
 \section{Benchmarking}
-To run benchmarks, we used Google's Cloud Compute product, because it allowed easy creation of machines with various configurations.
-Each benchmark was done on a virtual machine created to run the given benchmark.
-After the benchmark completed, the machines were destroyed.
-This allowed us to control the number of cores the JVM would use on every machine, since there is no good way to specify the number of cores the JVM will use for every thread it creates (including its garbage collection threads).
-A variety of small problems were implemented with serial recursive code, then invocations of the macros were added to the serial code.
-For any given test, the difference in the serial code and the parallel code is usually about 10 characters.
-I will present the results from a subset of the tests here.
-Each test was tested with a variety of different configurations and parallelism granularities where chosen to try and effectively use the resources available.
-For the tests with a fixed granularity (eg $n < 30$) the granularity was held constant for all tests.
-Some tests have dynamic granularity computed with respect to the number of cores available and the size of the input.
-For these tests, the granularity was not fixed.
+To run benchmarks, I used Google's Cloud Compute virtual machines.
+For each trial, a virtual machine was created.
+Each virtual machine had either $1$, $2$, $4$, or $6$ cores and $6$ gigabytes of RAM.
+First, the serial version of the code was run, then, on the same machine, the parallel version of the code was run.
+After both trials finished running, the data was copied back to my local machine and the virtual machine was destroyed.
+For every pair of serial/parallel executions, the speedup was computed.
+These per--machine speedups are used to generate the plots shown.
+
+To workaround the difficulties JVM benchmarking introduces, the Criterium\footnote{\url{https://github.com/hugoduncan/criterium}} library was used for Clojure code and the Caliper\footnote{\url{https://github.com/google/caliper}} library was used with Java code.
+The JVM does not expose a mechanism to control the total number of threads it creates (including garbage collection threads).
+Because each benchmark was run on it's own virtual machine with a constrained number of cores, the number of threads the JVM could create was controlled.
 
 \subsection{Fibonacci}
-First we will look at the classical recursive fibonacci example.
-This benchmark was implemented in both Clojure and Java, to compare the speedups gotten with the macros vs the speedup gotten with handwritten Java Fork/Join code.
-Figure \ref{fig:javafib} displays the average speedup over many tests for the Java Fork/Join code.
-As we can see, we only get about a 3x speedup with 6 cores, in the simple Java implementation.
+First we will look at the classical recursive Fibonacci example.
+Figure \ref{fig:fib_code} shows the code used for the serial and the parallel benchmarks.
+In Figure \ref{fig:fibparfun} the results from many trials of this code running $1$, $2$, $4$, and $6$ cores.
+Each benchmark computes \mintcode{(fib 35)}.
+We see that we get about a 3x speedup with $6$ cores.
+This speedup isn't quite what we would hope to see, but, as can be seen in Figure \ref{fig:javafib}, the handwritten Java Fork/Join implementation gets about the same speedup with $6$ cores on these virtual machines.
 
-<<javafib, fig.pos="H", fig.cap="Fibonacci Java Performance", echo=F>>=
+\begin{figure}
+\centering
+\begin{subfigure}[b]{0.5\textwidth}
+\centering
+\begin{minted}{clojure}
+(defn fib [n]
+  (if (or (= 0 n) (= 1 n))
+    1
+    (+
+     (fib (- n 1))
+     (fib (- n 2)))))
+\end{minted}
+\subcaption{Serial recursive Fibonacci}
+\end{subfigure}%
+\begin{subfigure}[b]{0.5\textwidth}
+\centering
+\begin{minted}{clojure}
+(defparfun fibparfun [n] (< n 35)
+  (if (or (= 0 n) (= 1 n))
+    1
+    (+
+     (fibparfun (- n 1))
+     (fibparfun (- n 2)))))
+\end{minted}
+\subcaption{Fibonacci with \defparfun{}}
+\end{subfigure}
+\caption{}
+\label{fig:fib_code}
+\end{figure}
+
+<<fibparfun, fig.pos="H", fig.cap="Fibonacci Clojure (defparfun) Performance", echo=F, out.width='0.7\\linewidth', fig.align='center'>>=
+cljfib <- plot_means_error_for_benchmark(data, "fib", "parfun")
+@
+
+<<javafib, fig.pos="H", fig.cap="Fibonacci Java Performance", echo=F, out.width='0.7\\linewidth', fig.align='center'>>=
 jfib <- plot_means_error_for_benchmark(data, "fib", "fj")
@@
 
-<<fibparfun, fig.pos="H", fig.cap="Fibonacci Clojure (defparfun) Performance", echo=F>>=
-cljfib <- plot_means_error_for_benchmark(data, "fib", "parfun")
+The large variance we see in the Clojure benchmarks is somewhat disturbing, especially since it does not show up in the Java results.
+Since the large variance does not appear in the Java benchmarks, it is probably not a result of variable performance in the Google Cloud Platform virtual machines.
+In Figure \ref{fig:fib_sd} I show the standard deviation of the mean runtime for the serial and parallel Clojure Fibonacci functions, along with their Java counterparts.
+Notice that the Java results do not suffer from the same extremely large variability problem.
+While I cannot completely explain the variability in the results, it seems to be caused by Clojure's function implementation, and the increased pressure that the Fork/Join wrapper tasks places on the JVM garbage collector.
+Every Clojure function is an \code{Object} created following the Clojure \code{IFn} interface\footnote{\url{http://clojure.org/reference/special_forms\#fn}}.
+When running on the \code{ForkJoinPool}, each function is further wrapped in a \code{RecursiveTask} object, causing additional allocations.
+We can avoid excessive task creation by controlling the granularity of parallelism, but, these tasks will always move functions into \code{RecursiveTask}s and introduce some amount of additional overhead.
+The large number of allocations causes additional garbage collector work.
+The garbage collector behaves somewhat non---deterministically, so I believe this is the explanation for the large variation in runtime for the serial and recursive Fibonacci code.
+The Fibonacci example highlights this problem because the function call overhead greatly exceeds the amount of work each call is doing.
+We will see an example for which this is not the case in Section \ref{label:id3_bench}.
+
+<<fib_sd, fig.pos="H", fig.cap=" ", echo=F, out.width='0.7\\linewidth', fig.align='center'>>=
+fib <- data[data$spec.name %like% "fib.big",]
+
+important <- c("cores", "mean.runtime")
+
+clj_serial <- ddply(fib[fib$spec.name %like% "serial",important], "cores", colwise(sd))
+clj_parfun <- ddply(fib[fib$spec.name %like% "parfun",important], "cores", colwise(sd))
+
+java_serial <-ddply(data[data$spec.name %like% "serial_java", important], "cores", colwise(sd))
+java_par    <-ddply(data[data$spec.name %like% "fj_java", important], "cores", colwise(sd))
+
+# d <- merge(clj_serial, clj_parfun, "cores", suffixes = c("serial", "parfun"))
+d <- join_all(list(clj_serial, clj_parfun, java_serial, java_par), by="cores")
+
+nnames <- c("Clojure serial", "Clojure parallel w/ defparfun", "java serial", "java fork/join")
+names(d)[2] <- nnames[1]
+names(d)[3] <- nnames[2]
+names(d)[4] <- nnames[3]
+names(d)[5] <- nnames[4]
+
+barplot(t(as.matrix(d[,nnames])), beside=TRUE, names=d$cores, legend=nnames,
+        main = "Fibonacci standard deviations", xlab="cores", ylab="mean runtime standard deviation")
 @
 
-<<id3parfun, fig.pos="H", fig.cap="id3 Clojure (defparfun) Performance", echo=F>>=
+\subsection{ID3} \label{label:id3_bench}
+I also implemented a simple ID3\footnote{\url{https://en.wikipedia.org/wiki/ID3_algorithm}} classifier in Clojure.
+The code is bit longer, so it is not included in this paper, but it can be found on this project's Github page\footnote{\url{https://github.com/dpzmick/auto_parallel}}.
+
+For each benchmark, a random 1,000 element dataset was created.
+Each element of the dataset was given 100 random features.
+The \code{ID3} algorithm implementation ran until it was out of attributes to pivot on.
+
+The \code{ID3} code does much more work in each function call, so the additional overhead created by Clojure's function implementation does not seem to impact the results as much as it does in the Fibonacci benchmark.
+
+Figure \ref{fig:id3parfun} shows that we get the 3x speedup we expect on these virtual machines with the \code{ID3} algorithm.
+
+<<id3parfun, fig.pos="H", fig.cap="id3 Clojure (defparfun) Performance", echo=F, out.width='0.7\\linewidth', fig.align='center'>>=
 id3 <- plot_means_error_for_benchmark(data, "id3", "parfun")
@@
 
-\section{Conclusions}
+\section{Conclusions and Future Work}
 % the jvm is (mostly) awesome

### 2016-04-19 paper/paper.Rnw
use the real data
paper

@@
 source("../vis/basic_plot.R")
-data <- read.csv("../newnew.csv")
+# data <- read.csv("../newnew.csv")
+data <- read.csv("../data.csv")
 @
@@
 % only discuss task based parallelism in this section
+Clojure has access to all of the JVM, so it has access to the Java \code{ForkJoinPool} library\cite{Lea2000}.
+The \code{ForkJoinPool} library allows Java programmers to create lightweight, recursive tasks, then execute them on an efficient, work stealing, \code{ExecutionEngine}.
+Users of a \code{ForkJoinPool} create a subclass of \code{RecursiveTask} to compute the value of some function.
+Each of these tasks may create more \code{RecursiveTask}s, submit them to the \code{ForkJoinPool}, then wait for their subtasks to complete, without blocking the pool.
+The execution engine for the \code{ForkJoinPool} is a thread pool.
+Each thread maintains it's own queue of tasks to work on.
+Whenever a thread runs out of tasks to work on, it steals tasks from other threads.
+Work stealing has been used a scheduling mechanism is a variety of modern threading libraries and has been very successful \cite{Lea2000, Blumofe1994, Blumofe1995a}
+Clojure programmers can use the \code{ForkJoinPool} from Clojure, but the interface isn't exactly programmer friendly.
 
-\subsection{Clojure and other lisps}
-% lparallel
-% futures
-% claypoole futures
-Makes lots of threads, unatural blah.
-Note sure what's bad about lparallel other than it isn't implemented in Clojure.
-Clojure also has access to Java's libraries, like ForkJoin.... but hard to use and not really a part of the Clojure language.
+Clojure also has some built in support for task parallelism via \code{future}\footnote{\url{https://clojuredocs.org/clojure.core/future}}.
+Each call to future creates a thread to compute the result of a function call in the background.
+When a user is ready to access a value, they can \code{deref} the future to get the computed value (\code{deref} will block until the value is computed).
+For a user, \code{future}s initially seem like a fairly natural way to parallelize recursive functions.
+Unfortunately, Futures only work well when a small number of tasks are created or when the tasks do a lot of blocking I/O.
+Thread creation overhead is high, and people users must be careful not to create an excessive number of threads or create threads to do too little work.
+Claypoole offers many improvements to the builtin Clojure \code{future}s, but it still creates a thread per task.
+The work in this paper is conceptually similar, but the interface I have used differs dramatically from the \code{future}s interface.
+My code also creates tasks for every thread, but these tasks are created on a Java \code{ForkJoinPool}, so the tasks are much more lightweight.
 
-Briefly, a \code{ForkJoinPool} is an executor for lightweight (often recursive) tasks.
-The execution engine uses a work stealing scheduler to schedule the lightweight tasks across a thread pool.
+In Common Lisp, the \code{lparallel}\footnote{\url{https://lparallel.org/overview/}} library and macros are available.
+The macros I have implemented are very similar to the macros \code{lparallel} provides, but, Common Lisp programming conventions are not as ideal for these kinds of macros.
 
 \subsection{Other languages}
-% - cilk/tbb/openmp
-Compiler support is bad.
-Performance is quite good.
-Only real downside is that these are not implemented in Clojure.
-ForkJoin pools are essentially Cilk \cite{CilkReference, Frigo1998}.
-Both provide the support and runtime needed to easily parallelize code, but neither give programmers the ability to mark a function as ``parallel'' then let the macro system deal with granularity and the (small) syntactic overhead (cilk has very little extra syntax).
+In other languages, frameworks like Cilk++\footnote{\url{https://www.cilkplus.org/}}, OpenMP\footnote{\url{http://openmp.org/wp/}}, Threading Building Blocks\footnote{\url{https://www.threadingbuildingblocks.org/}} all exist.
+Some of these require compiler support and advanced dependence analysis to guarantee correctness.
+These libraries all have great performance, but they can't be leveraged from Clojure.
+The macros I've implemented are an attempt to bring some of the niceties of these libraries and compiler extensions to Clojure, without the need to compiler support or complicated dependence analysis.
 
@@
 Before discussing the macros I've implemented, I need to loosely define a ``mostly pure'' function.
-A mostly pure function is a function that has no side effects which directly impact the values of other user-defined values, at the current level of abstraction.
+A mostly pure function is a thread---safe function with no side effects directly impacting the values of user-defined values, at the current level of abstraction.
 Mostly pure functions can be reordered (or interleaved) without impacting the values of user variables, although the change may impact I/O behavior and output order of a program.
@@
 For example, it may not matter if we reorder \mintcode{(download file1)} and \mintcode{(download file2)} for a programmer writing a web scraper, but it may matter to a programmer writing a I/O constrained server.
-When a programmer feels that it may be acceptable to change the order of execution of calls to mostly pure functions (the call is ``portable''), we can reorder them subject to these constraints:
+When a programmer feels that it may be acceptable to change the order of, or interleave, calls to mostly pure functions (the call is ``portable''), we can reorder them subject to these constraints:
 
@@
     \node (start)       [badnode, below of=start0]                       {\mintcode{(if (pred? a b c ...)}};
-    \node (iftruedo)    [badnode, below of=start, right of=start]        {\mintcode{(do}};
+    \node (iftruedo)    [goodnode, below of=start, right of=start]       {\mintcode{(do}};
     \node (iftruelet1)  [goodnode, below of=iftruedo, right of=iftruedo] {\mintcode{;; some code that uses a, b, and/or c}};
@@
 Many Clojure functions fit this definition due to Clojure programming conventions and default immutable data structures.
-This concept is not rigorously defined as there is not a good definition for the semantics of such functions. % TODO find references
-Instead, in this paper, we rely on the judgment of the programmer to tell us when a mostly pure function is portable.
+In this paper, we rely on the judgment of the programmer to tell us when a mostly pure function is portable.
 
@@
 \parlet{} is a modified version of the Clojure \code{let} form.
-The \parlet{} macro has exactly the same semantics as Clojure's \code{let}, but it evaluates all of its bindings in parallel.
+The \parlet{} macro has exactly the same behavior as Clojure's \code{let}, but it evaluates all of its bindings in parallel.
 For example, suppose I had some long running function \code{foo}.
@@
 In this example, the expressions \mintcode{(foo value1)} and \mintcode{(foo value2)} are both evaluated as \code{ForkJoinTask}s in a \code{ForkJoinPool}\cite{Lea2000}.
-
-The calls to \code{foo} are both \code{forked} immediately, then we attempt to evaluate the body of the \code{let}.
+The calls to \code{foo} are both forked immediately, then we attempt to evaluate the body of the \parlet{}.
 This means that the code in the body of the \code{let} which does not depend on the computations of \code{a} and \code{b} can execute without delay.
-
-Since the \code{ForkJoinPool} is designed for recursive workloads, it allows tasks which are currently executing to create new tasks, submit them to the pool, then wait for the task to complete.
-None of the pool's threads will block when this occurs.
-This means that nested \parlet{} forms (Figure \ref{fig:parlet_noblock}) will execute without blocking.
+Additionally, since the \code{ForkJoinPool} is designed for recursive workloads, tasks which are currently executing can create new tasks, submit them to the pool, then wait for the task to complete, without blocking execution.
+This means that nested \parlet{} forms (Figure \ref{fig:parlet_noblock}) will not block each other.
 
@@
 \end{figure}
-
-In Figure \ref{fig:parlet_noblock}, calls to the function \code{foo} will be processed in the background until we reach the line using the variables that the results are bound to.
-This means that parlets can be nested with little concern (again as long as functions are pure).
 
@@
 
-This transformation is only safe when \code{foo} (and more generally, and function call in the bindings) is a mostly pure function.
-Here, we punt the definition of mostly pure to the programmer.
+\subsection{Correctness}
+This transformation is only safe when \code{foo} (and more generally, any function call in the bindings) is a mostly pure function.
 If the programmer chooses to use a \parlet{} form, we assume that the functions are mostly pure.

### 2016-04-19 paper/paper.Rnw
remove stuff im not using

@@
-%\documentclass[12pt, letterpapper]{proc}
-\documentclass[12pt, letterpapper, titlepage]{article}
+\documentclass[10pt, letterpapper]{proc}
+%\documentclass[12pt, letterpapper, titlepage]{article}
 \usepackage{hyperref}
@@
 \usemintedstyle{borland}
-
-\usepackage{subcaption}
 
@@
 
-Clojure is a lisp--like language running on the Java Virtual Machine.
-The language's default immutable data structures and sophisticated Software Transactional Memory (STM) system make it well suited for parallel programming.
+Clojure is a lisp--like language running on the Java Virtual Machine.\cite{Hickey2008}
+The language's default immutable data structures and sophisticated Software Transactional Memory (STM) system make it well suited for parallel programming.\cite{Kraus2009}
 Because Clojure runs on the Java Virtual Machine, Clojure developers can take advantage of existing cross--platform parallelism libraries, such as Java's excellent \code{ExecutorService} framework, to write parallel code.
@@
 However, taking advantage of Clojure's parallel potential is not entirely straightforward.
-STM has proven to be very successful as a clear construct for concurrent programming\cite{Jones2007a}, but these constructs are often too low level to be of much use to developers whose central concerns are not parallelism\cite{Boehm2009}.
+STM has proven to be very successful construct for concurrent programming\cite{Jones2007a}, but these constructs are often too low level to be of much use to developers whose central concerns are not parallelism\cite{Boehm2009}.
 
 As a result, there are a variety of libraries designed to allow developers to take advantage of the parallelism potential in Clojure.
-Clojure builtins such as pmap\footnotemark[1] and the new reducers\footnotemark[2] library provide data parallel sequence manipulation and transformation functions.
+Clojure builtins such as pmap\footnotemark[1] and reducers\footnotemark[2] library provide data parallel sequence manipulation and transformation functions.
 Third---party libraries like Tesser\footnotemark[3] and Claypoole\footnotemark[4] provide more data parallel APIs with slightly different goals than the builtin functions.
@@
 
-\section{Mostly Pure Functions}
+\section{Mostly Pure Functions} \label{label:mostly_pure}
 Before discussing the macros I've implemented, I need to loosely define a ``mostly pure'' function.
@@
 
-\tikzstyle{goodnode} = [rectangle, draw=blue, text width=8.8cm]
-\tikzstyle{badnode}  = [rectangle, draw=none,  text width=8.8cm]
+\tikzstyle{goodnode} = [rectangle, draw=blue, text width=0.8\linewidth]
+\tikzstyle{badnode}  = [rectangle, draw=none,  text width=0.8\linewidth]
 
-\begin{figure}[H]
+\begin{figure}[h]
 \centering
-\begin{tikzpicture}[node distance=0.6cm,auto]
+\begin{tikzpicture}[node distance=0.62cm,auto]
     \node[coordinate] (0) at (0.7,0) {};
-    \node (start0)      [badnode]                                        {\mintcode{;; we cannot move the call outside of the if.} \\ \mintcode{;; There is a path on which the call will never occur}};
+    \node (start0)      [badnode]                                        {\mintcode{;; cannot move the call outside}};
     \node (start)       [badnode, below of=start0]                       {\mintcode{(if (pred? a b c ...)}};
     \node (iftruedo)    [goodnode, below of=start, right of=start]       {\mintcode{(do}};
-    \node (iftruelet1)  [goodnode, below of=iftruedo, right of=iftruedo] {\mintcode{;; some code that uses a, b, and/or c}};
+    \node (iftruelet1)  [goodnode, below of=iftruedo, right of=iftruedo] {\mintcode{;; code that uses a, b, and/or c}};
     \node (iftruelet2)  [goodnode, below of=iftruelet1]                  {\mintcode{(foo a b c))}};
@@
 
-\begin{figure}[H]
+\begin{figure}[h]
 \centering
-\begin{tikzpicture}[node distance=0.6cm,auto]
+\begin{tikzpicture}[node distance=0.61cm,auto]
     \node[coordinate] (0) at (1.0,0) {};
@@
     \node (a)           [badnode, below of=start, right of=start] {\mintcode{(let [a (bar 10)]}};
-    \node (b)           [goodnode, below of=a, right of=a]        {\mintcode{;; some code that uses a}};
+    \node (b)           [goodnode, below of=a, right of=a]        {\mintcode{;; code that uses a}};
     \node (c)           [goodnode, below of=b]                    {\mintcode{(let [b (bar a)]}};
-    \node (d)           [goodnode, below of=c, right of=c]        {\mintcode{;; some code that uses a and/or b}};
+    \node (d)           [goodnode, below of=c, right of=c]        {\mintcode{;; code that uses a and/or b}};
     \node (e)           [goodnode, below of=d]                    {\mintcode{(foo a)))}};
@@
 This transformation is only safe when \code{foo} (and more generally, any function call in the bindings) is a mostly pure function.
-If the programmer chooses to use a \parlet{} form, we assume that the functions are mostly pure.
+If the programmer chooses to use a \parlet{} form, we assume that the functions called in the bindings are mostly pure.
 This simple dependency check, as well as the programmers promise that all function called are mostly pure Clojure code, allow us to ensure correct parallelism with this macro.
 
-\section{parexpr}
+% \section{parexpr}
 
-The \parexpr{} macro breaks down expressions and (aggressively) evaluates them in parallel.
-Suppose again that I had a long running function \code{foo} which I wanted to call twice, and add the results.
-The code in Figure \ref{fig:parexpr} will make both of the long running calls to \code{foo} in parallel.
-The \parexpr{} macro crawls the expression and expands it into multiple nested \parlet{} forms, filling each \parlet{} with as many evaluations as it can.
+% The \parexpr{} macro breaks down expressions and (aggressively) evaluates them in parallel.
+% Suppose again that I had a long running function \code{foo} which I wanted to call twice, and add the results.
+% The code in Figure \ref{fig:parexpr} will make both of the long running calls to \code{foo} in parallel.
+% The \parexpr{} macro crawls the expression and expands it into multiple nested \parlet{} forms, filling each \parlet{} with as many evaluations as it can.
 
-\begin{figure}[h]
-\centering
-\RecustomVerbatimEnvironment{Verbatim}{BVerbatim}{}
-\begin{minted}{clojure}
-(parexp (+ (foo value1) (foo value2)))
-\end{minted}
-\caption{Example using parexpr to evaluate long running functions}
-\label{fig:parexpr}
-\end{figure}
+% \begin{figure}[h]
+% \centering
+% \RecustomVerbatimEnvironment{Verbatim}{BVerbatim}{}
+% \begin{minted}{clojure}
+% (parexp (+ (foo value1) (foo value2)))
+% \end{minted}
+% \caption{Example using parexpr to evaluate long running functions}
+% \label{fig:parexpr}
+% \end{figure}
 
@@
 
-The \defparfun{} macro is the most interesting of the 3 macros.
 \defparfun{} allows a programmer to parallelize calls to a recursive function.
+The \defparfun{} macro supports a granularity argument, allowing the programmer to specify when they would like to stop creating additional parallel tasks.
+The macro emits the expression provided for granularity inside of an \code{if} statement at the top of the function, so the programmer can use any arbitrary condition, including conditions dependent on the functions arguments, to decide when to stop spawning new tasks.
+
 For an example see Figure \ref{fig:fib_code}.
-This defines a parallel Fibonacci function, which will only execute in parallel when the value of it's argument is greater than $30$.
+This defines a parallel Fibonacci function, which will only execute in parallel when the value of it's argument is greater than $35$.
 
-\subsection{Imposed Constraints} \label{label:mpure_algo}
-The objective of the \defparfun{} macro is to enable simple, predictable parallelism for recursive functions which are mostly pure.
-We also aim to support flexible functions, but we cannot promise correctness when the macro is used with flexible functions.
+\begin{figure}[h]
+\RecustomVerbatimEnvironment{Verbatim}{BVerbatim}{}
+\centering
+\begin{minted}{clojure}
+(defparfun fib [n] (< n 35)
+  (if (or (= 0 n) (= 1 n))
+    1
+    (+
+     (fib (- n 1))
+     (fib (- n 2)))))
+\end{minted}
+\caption{Fibonacci function with \defparfun{} added}
+\label{fig:fib_code}
+\end{figure}
 
-A mostly pure function call may be moved to a node $P$ in the control flow graph for the function iff every path through $P$ in the original function would execute the function call, and if the arguments to the function call are all available in $P$.
-We do not allow function calls to be introduced on a path which they would not have existed on before to preserve
+\subsection{Implementation} \label{label:mpure_algo}
+To do implement this macro, I first expand all of the \code{let} expressions in the function so that every \code{let} only has a single binding.
+Then, the syntax tree is crawled, finding every recursive function call.
+Every recursive call is replaced by a newly introduced variable.
+At the termination of each recursive call, the bindings introduced in the subexpression are returned.
+Once all of the subexpressions are evaluated, the algorithms decides if it must emit a \code{let} form for any of the bindings the subexpressions introduced.
+If the expression in question is an \code{if} statement, a let statement is introduced for the \code{true} branch and on the \code{false} branch, each containing all of the bindings introduced by the subexpressions for each branch.
+If the statement is a let statement, only the bindings which depend on the binding introduced by the let statement are emitted (following the let of course).
+This operation effectively pushed the evaluation of the function all as far ``up'' in the function as it can, following the constraints described in Section \ref{label:mostly_pure}.
 
-% we can move these functions around as long as all of their arguments are not changed
-% - the output order of the program may change slightly, but the results will not
-% don't move all calls to the top because don't want to create unpredictable performance or do wasteful evaluations, or break dependencies.
+After this transformation, it is possible to replace the \code{let} forms which bind the function results to their values with \parlet{} forms providing the same bindings.
+The introduction of the \parlet{} form introduces parallelism, so each recursive call will execute in the \code{ForkJoin} pool, in parallel.
 
-\subsection{Moving portable mostly pure functions}
-The macro first identifies any recursive call sites.
-It then introduces a name for the call, then pushes the evaluation of the function all as far ``up'' in the function as it can.
-We stop pushing a function call higher in a function when we encounter an \code{if} form, or we encounter any existing \code{let} expression which introduces a value that the function call depends on.
-We can think of this operation as an operation on a single ``basic block'' in the function.
-The function call can be freely moved inside of the function as long as the function call does not introduce any new behavior to the function.
+Because the transformation does not violate any of the properties defined for mostly pure functions, this transformation is safe when the function being declared as a \code{parfun} is mostly pure.
 
-After this transformation, it is possible to replace the let forms which bind the function results to their values with \parlet{} forms providing the same bindings.
-The introduction of the parlet form introduces parallelism, so each recursive call will execute in the \code{ForkJoin} pool, in parallel.
-Each of these subsequent recursive calls may create more tasks, which will also be executed in the pool.
-\code{ForkJoin} pools are designed to handle this type of computation, so the computation will proceed without blocking (as long as function is pure and performs no thread--blocking I/O).
+\begin{figure}[h]
+\RecustomVerbatimEnvironment{Verbatim}{BVerbatim}{}
+\centering
+\begin{minted}{clojure}
+(defn fib [n]
+ ;; granularity check
+ (if (< n 35)
+  (if (or (= 0 n) (= 1 n))
+   1
+   (+ (fib (- n 1))
+      (fib (- n 2))))
+
+  ;; recursive case
+  (if (or (= 0 n) (= 1 n))
+   1
+   (parlet [expr17300 (fib (- n 1))
+            expr17301 (fib (- n 2))]
+    (+ expr17300 expr17301)))))
+\end{minted}
+\caption{Transformed function}
+\label{fig:parfun_result}
+\end{figure}
 
@@
 \subsection{Fibonacci}
-First we will look at the classical recursive Fibonacci example.
-Figure \ref{fig:fib_code} shows the code used for the serial and the parallel benchmarks.
-In Figure \ref{fig:fibparfun} the results from many trials of this code running $1$, $2$, $4$, and $6$ cores.
-Each benchmark computes \mintcode{(fib 35)}.
-We see that we get about a 3x speedup with $6$ cores.
-This speedup isn't quite what we would hope to see, but, as can be seen in Figure \ref{fig:javafib}, the handwritten Java Fork/Join implementation gets about the same speedup with $6$ cores on these virtual machines.
 
-\begin{figure}
-\centering
-\begin{subfigure}[b]{0.5\textwidth}
+\begin{figure}[h]
+\RecustomVerbatimEnvironment{Verbatim}{BVerbatim}{}
 \centering
 \begin{minted}{clojure}
-(defn fib [n]
+(defun fib [n]
   (if (or (= 0 n) (= 1 n))
@@
 \end{minted}
-\subcaption{Serial recursive Fibonacci}
-\end{subfigure}%
-\begin{subfigure}[b]{0.5\textwidth}
-\centering
-\begin{minted}{clojure}
-(defparfun fibparfun [n] (< n 35)
-  (if (or (= 0 n) (= 1 n))
-    1
-    (+
-     (fibparfun (- n 1))
-     (fibparfun (- n 2)))))
-\end{minted}
-\subcaption{Fibonacci with \defparfun{}}
-\end{subfigure}
-\caption{}
-\label{fig:fib_code}
+\caption{Serial Fibonacci function}
+\label{fig:fib_serial}
 \end{figure}
 
-<<fibparfun, fig.pos="H", fig.cap="Fibonacci Clojure (defparfun) Performance", echo=F, out.width='0.7\\linewidth', fig.align='center'>>=
+First we will look at the classical recursive Fibonacci example.
+Figure \ref{fig:fib_serial} shows the serial benchmark code; \ref{fig:fib_code} shows the code parallel benchmark code.
+In Figure \ref{fig:fibparfun} the results from many trials of this code running $1$, $2$, $4$, and $6$ cores.
+Each benchmark computes \mintcode{(fib 39)}.
+We see that we get about a 3x speedup with $6$ cores.
+This speedup isn't quite what we would hope to see, but, as can be seen in Figure \ref{fig:javafib}, the handwritten Java \code{ForkJoinPool} implementation gets about the same speedup with $6$ cores on these virtual machines.
+Previous \code{ForkJoinPool} benchmarks have shown much better speedups for similar code\cite{Lea2000}, so I suspect that the virtual machine configuration is somewhat responsible for the discrepancy in results.
+
+<<fibparfun, fig.pos="H", fig.cap="Fibonacci Clojure (defparfun) Performance", echo=F, out.width='\\linewidth', fig.align='center'>>=
 cljfib <- plot_means_error_for_benchmark(data, "fib", "parfun")
@@
 
-<<javafib, fig.pos="H", fig.cap="Fibonacci Java Performance", echo=F, out.width='0.7\\linewidth', fig.align='center'>>=
+<<javafib, fig.pos="H", fig.cap="Fibonacci Java Performance", echo=F, out.width='\\linewidth', fig.align='center'>>=
 jfib <- plot_means_error_for_benchmark(data, "fib", "fj")
@@
 The large variance we see in the Clojure benchmarks is somewhat disturbing, especially since it does not show up in the Java results.
-Since the large variance does not appear in the Java benchmarks, it is probably not a result of variable performance in the Google Cloud Platform virtual machines.
+Since it does not appear in the Java benchmarks, it is not a result of variable performance in the Google Cloud Platform virtual machines.
 In Figure \ref{fig:fib_sd} I show the standard deviation of the mean runtime for the serial and parallel Clojure Fibonacci functions, along with their Java counterparts.
-Notice that the Java results do not suffer from the same extremely large variability problem.
-While I cannot completely explain the variability in the results, it seems to be caused by Clojure's function implementation, and the increased pressure that the Fork/Join wrapper tasks places on the JVM garbage collector.
+Notice that the Java results do not have nearly as high deviations from mean runtime.
+While I cannot completely explain the variability, it seems to be caused by the increased pressure Clojure's function implementation and the \code{ForkJoinPool} wrapper tasks places on the JVM garbage collector.
 Every Clojure function is an \code{Object} created following the Clojure \code{IFn} interface\footnote{\url{http://clojure.org/reference/special_forms\#fn}}.
 When running on the \code{ForkJoinPool}, each function is further wrapped in a \code{RecursiveTask} object, causing additional allocations.
-We can avoid excessive task creation by controlling the granularity of parallelism, but, these tasks will always move functions into \code{RecursiveTask}s and introduce some amount of additional overhead.
 The large number of allocations causes additional garbage collector work.
 The garbage collector behaves somewhat non---deterministically, so I believe this is the explanation for the large variation in runtime for the serial and recursive Fibonacci code.
+We can avoid excessive task creation by controlling the granularity of parallelism, to an extent.
 The Fibonacci example highlights this problem because the function call overhead greatly exceeds the amount of work each call is doing.
@@
 
-<<fib_sd, fig.pos="H", fig.cap=" ", echo=F, out.width='0.7\\linewidth', fig.align='center'>>=
+<<fib_sd, fig.pos="H", fig.cap=" ", echo=F, out.width='\\linewidth', fig.align='center'>>=
 fib <- data[data$spec.name %like% "fib.big",]
@@
 
-<<id3parfun, fig.pos="H", fig.cap="id3 Clojure (defparfun) Performance", echo=F, out.width='0.7\\linewidth', fig.align='center'>>=
+<<id3parfun, fig.pos="H", fig.cap="id3 Clojure (defparfun) Performance", echo=F, out.width='\\linewidth', fig.align='center'>>=
 id3 <- plot_means_error_for_benchmark(data, "id3", "parfun")
@@
 \section{Conclusions and Future Work}
-% the jvm is (mostly) awesome
-% immutability makes things like this really straightforward
+Clojure's conventions make transformations like this possible, and simple.
 

### 2016-04-19 paper/paper.Rnw
updated paper

@@
 When running on the \code{ForkJoinPool}, each function is further wrapped in a \code{RecursiveTask} object, causing additional allocations.
-The large number of allocations causes additional garbage collector work.
+This effectively moves the stack for the recursive function to the heap.
+There is so upside to this; it eliminates the stack depth limit (which is an issue in Clojure because Clojure cannot implement tail call optimization), the large number of allocations creates a large amount garbage collector pressure.
 The garbage collector behaves somewhat non---deterministically, so I believe this is the explanation for the large variation in runtime for the serial and recursive Fibonacci code.
@@
 \section{Conclusions and Future Work}
-Clojure's conventions make transformations like this possible, and simple.
+The Clojure macros I've implemented perform transformation which can speedup Clojure code to a degree which matches the speedups attained using handwritten Java code, running on the same hardware.
+Parallelism is difficult, and automatic parallelism is possible\cite{Banerjee1993}, but these techniques are complicated and often do not get the desired results and the research community has begun to feel the need for explicit parallelism in programs.\cite{Arvind2010}
+Languages like Clojure are well suited for this parallelism and techniques like mine can easily be implemented.
+In a language with a strong STM system and immutable structures, such transformations are easy to reason about, making it much simpler for programmers to implement explicitly parallel programs.
+
+Macros of this style do not inhibit the programmers ability to use the other mechanisms implemented in the language, although interoperability with them could be improved.
+For example, one of the tests which was not discussed in this paper involved using the STM system from within a function declared with \defparfun{}.
+Benchmarks on this code behaved correctly and performance improved as expected.
+However, if a programmer attempted to use a \code{pmap} or \code{future} inside of a \defparfun{} or \parlet{}, the two systems would create separate thread pools and the number of created threads would be large, possibly causing poor performance.
+There are also a variety of other useful macros in lparallel\footnote{\url{https://lparallel.org/}} that may be useful to implement in Clojure and would complement the macros I've implemented in this project.
+
+It would also be useful to use profiling tools and implement static analysis tools which would detect potential locations for these macros.
+These tools could be used by developers to look for potential opportunities for parallelism, and they could be used to demonstrate the claim that many Clojure programs may benefit from use of these macros.
 

### 2016-04-19 paper/paper.Rnw
paper

@@
 \usemintedstyle{borland}
+
+\usepackage{subcaption}
 
@@
 
-Clojure is a lisp--like language running on the Java Virtual Machine.\cite{Hickey2008}
-The language's default immutable data structures and sophisticated Software Transactional Memory (STM) system make it well suited for parallel programming.\cite{Kraus2009}
+Clojure is a lisp--like language running on the Java Virtual Machine\cite{Hickey2008}.
+The language's default immutable data structures and sophisticated Software Transactional Memory (STM) system make it well suited for parallel programming\cite{Kraus2009}.
 Because Clojure runs on the Java Virtual Machine, Clojure developers can take advantage of existing cross--platform parallelism libraries, such as Java's excellent \code{ExecutorService} framework, to write parallel code.
@@
 As a result, there are a variety of libraries designed to allow developers to take advantage of the parallelism potential in Clojure.
-Clojure builtins such as pmap\footnotemark[1] and reducers\footnotemark[2] library provide data parallel sequence manipulation and transformation functions.
+Clojure built ins such as \code{pmap}\footnotemark[1] and reducers\footnotemark[2] library provide data parallel sequence manipulation and transformation functions.
 Third---party libraries like Tesser\footnotemark[3] and Claypoole\footnotemark[4] provide more data parallel APIs with slightly different goals than the builtin functions.
 Developers have a good relationship with data parallel problems\cite{Okur2012a}, but Clojure's nature as a functional language with immutable structures also makes it possible to easily exploit control parallelism (also known as task parallelism\cite{Andradea, Rodr}).
-An astute reader may observe that both Claypoole and the Clojure standard library include task parallel functions.
-These will be discussed in the Section \ref{label:related_work}
 
@@
 
-Using Clojure's macro system, I have implemented a set of macros which allow developers to take advantage of Clojure's parallelism potential when their existing code is written such that parallelism is exposed through control flow.
-I have shown that it is possible to attain reasonable degrees of parallelism with minimal code changes, with respect to serial code, using these macros.
-The intended users of these macros are developers whose primary concern is not performance, but may benefit from a simple mechanism with which they can take advantage of the many cores in their machines.
-Developers who are extremely concerned with performance and want a high degree of control should turn elsewhere (perhaps even to Java) to write their highly tuned code.
+Using Clojure's macro system, I have implemented a set of macros which allow developers to take advantage of Clojure's task parallelism potential.
+I have shown that it is possible to attain reasonable degrees of parallelism with minimal modification to existing serial code.
 
@@
 Clojure has access to all of the JVM, so it has access to the Java \code{ForkJoinPool} library\cite{Lea2000}.
-The \code{ForkJoinPool} library allows Java programmers to create lightweight, recursive tasks, then execute them on an efficient, work stealing, \code{ExecutionEngine}.
+The \code{ForkJoinPool} library allows Java programmers to create lightweight, recursive tasks which are executed by a thread pool.
+Each thread in the pool maintains a queue of tasks to work on.
+Whenever a thread runs out of tasks to work on, it steals tasks from other threads.
+Work stealing has been used a scheduling mechanism is a variety of modern threading libraries and has been very successful \cite{Lea2000, Blumofe1994, Blumofe1995a}
 Users of a \code{ForkJoinPool} create a subclass of \code{RecursiveTask} to compute the value of some function.
 Each of these tasks may create more \code{RecursiveTask}s, submit them to the \code{ForkJoinPool}, then wait for their subtasks to complete, without blocking the pool.
-The execution engine for the \code{ForkJoinPool} is a thread pool.
-Each thread maintains it's own queue of tasks to work on.
-Whenever a thread runs out of tasks to work on, it steals tasks from other threads.
-Work stealing has been used a scheduling mechanism is a variety of modern threading libraries and has been very successful \cite{Lea2000, Blumofe1994, Blumofe1995a}
-Clojure programmers can use the \code{ForkJoinPool} from Clojure, but the interface isn't exactly programmer friendly.
+Clojure programmers can use the \code{ForkJoinPool} libraries from Clojure, but the interface isn't exactly programmer friendly (or idiomatic).
 
-Clojure also has some built in support for task parallelism via \code{future}\footnote{\url{https://clojuredocs.org/clojure.core/future}}.
-Each call to future creates a thread to compute the result of a function call in the background.
-When a user is ready to access a value, they can \code{deref} the future to get the computed value (\code{deref} will block until the value is computed).
-For a user, \code{future}s initially seem like a fairly natural way to parallelize recursive functions.
-Unfortunately, Futures only work well when a small number of tasks are created or when the tasks do a lot of blocking I/O.
-Thread creation overhead is high, and people users must be careful not to create an excessive number of threads or create threads to do too little work.
-Claypoole offers many improvements to the builtin Clojure \code{future}s, but it still creates a thread per task.
+Clojure also has built in support for task parallelism via \code{future}\footnote{\url{https://clojuredocs.org/clojure.core/future}}.
+Each call to \code{future} spawns a thread to compute the result of a function call in the background.
+When a user is ready to access the computed value, they can \code{deref} the future to it.
+\code{deref} will block until the value is computed.
+For a user, \code{future} initially seem like a fairly natural way to parallelize recursive functions.
+Unfortunately, \code{future} only works well when a small number of tasks are created, or when the tasks do a lot of blocking I/O.
+Because thread creation overhead is high, users must be careful not to create an excessive number of threads or create threads which do too little work.
+Claypoole's implementation of \code{future} offers many improvements to the built in \code{future}.
+Their future implementation executes tasks created with \code{future} in a fixed sized thread pool, but, the \code{future} task will block the thread executing it if the task creates a new task, then immediately \code{deref}s it.
+This means that a programmer must be very careful not to create tasks in a Claypoole thread pool which depend on other tasks, otherwise their code may deadlock.
 The work in this paper is conceptually similar, but the interface I have used differs dramatically from the \code{future}s interface.
-My code also creates tasks for every thread, but these tasks are created on a Java \code{ForkJoinPool}, so the tasks are much more lightweight.
+My code also creates many tasks, but these tasks are created on a Java \code{ForkJoinPool}, so the tasks are much more lightweight, and they can be composed recursively.
 
@@
 \subsection{Other languages}
-In other languages, frameworks like Cilk++\footnote{\url{https://www.cilkplus.org/}}, OpenMP\footnote{\url{http://openmp.org/wp/}}, Threading Building Blocks\footnote{\url{https://www.threadingbuildingblocks.org/}} all exist.
+In other languages, frameworks like Cilk++\footnote{\url{https://www.cilkplus.org/}}, OpenMP\footnote{\url{http://openmp.org/wp/}}, Threading Building Blocks\footnote{\url{https://www.threadingbuildingblocks.org/}} have been implemented.
 Some of these require compiler support and advanced dependence analysis to guarantee correctness.
-These libraries all have great performance, but they can't be leveraged from Clojure.
 The macros I've implemented are an attempt to bring some of the niceties of these libraries and compiler extensions to Clojure, without the need to compiler support or complicated dependence analysis.
@@
 For example, it may not matter if we reorder \mintcode{(download file1)} and \mintcode{(download file2)} for a programmer writing a web scraper, but it may matter to a programmer writing a I/O constrained server.
-When a programmer feels that it may be acceptable to change the order of, or interleave, calls to mostly pure functions (the call is ``portable''), we can reorder them subject to these constraints:
+When a programmer feels that it may be acceptable to change the order of, or interleave, calls to mostly pure functions, we can reorder them subject to these constraints:
 
@@
 Many Clojure functions fit this definition due to Clojure programming conventions and default immutable data structures.
-In this paper, we rely on the judgment of the programmer to tell us when a mostly pure function is portable.
+In this paper, we rely on the judgment of the programmer to tell us when a function can be considered mostly pure.
 
@@
 % \code{const} functions promise not to modify any members of the object on which they are called (unless the member was declared \code{mutable}), essentially promising that the object will not change, from the user of the object's perspective, when the \code{const} function is called.
+
+\begin{figure*}[t!]
+\begin{subfigure}[b]{0.5\textwidth}
+\centering
+\RecustomVerbatimEnvironment{Verbatim}{BVerbatim}{}
+\begin{minted}{clojure}
+(parlet
+  [a (foo value1)
+   b (foo value2)]
+  ;; some other code here
+  (+ a b))
+\end{minted}
+\subcaption{Example of a parlet form}
+\label{fig:parlet_basic}
+\end{subfigure}%
+\begin{subfigure}[b]{0.5\textwidth}
+\centering
+\RecustomVerbatimEnvironment{Verbatim}{BVerbatim}{}
+\begin{minted}{clojure}
+(let
+  [a (fork (new-task (fn [] (foo 1))))
+   b (fork (new-task (fn [] (foo 2))))]
+  ;; some other code here
+  (+ (join a) (join b)))
+\end{minted}
+\subcaption{An expanded parlet form}
+\label{fig:parlet_basic_expanded}
+\end{subfigure}
+\caption{}
+\label{fig:parlet_example}
+\end{figure*}
+
+\newpage
 
@@
 I need to add the result of two calls to this function.
-In Figure \ref{fig:parlet_basic}, we use \parlet{} to make two calls to \code{foo}, then add the results.
+In Figure \ref{fig:parlet_example}, we use \parlet{} to make two calls to \code{foo}, then add the results.
+
+In this example, the expressions \mintcode{(foo value1)} and \mintcode{(foo value2)} are both evaluated as \code{RecursiveTask}s in a \code{ForkJoinPool}.
+The calls to \code{foo} are both forked immediately, then we attempt to evaluate the body of the \parlet{}.
+Each use of \code{a} and \code{b} is replaced with a call to \code{join}, to get the computed value.
+This means that the code in the body of the \code{let} which does not depend on the computations of \code{a} and \code{b} can execute without delay.
+Additionally, since the \code{ForkJoinPool} is designed for recursive workloads, tasks which are currently executing can create new tasks, submit them to the pool, then wait for the task to complete, without blocking tasks created by other \parlet{} calls.
+So, \parlet{} forms can be nested; a programmer does not have to worry if functions called in the bindings of a \parlet{} form also use \parlet{}.
+
+\subsection{Dependencies}
+The \parlet{} macro also supports simple dependency detection.
+Clojure \code{let} forms can use names defined previously in the same \code{let}.
+The bindings are evaluated from first to last.
 
@@
 \begin{minted}{clojure}
-(parlet
-  [a (foo value1)
-   b (foo value2)]
-   ... ; some other code here
-  (+ a b))
-\end{minted}
-\caption{Example of a parlet form}
-\label{fig:parlet_basic}
-\end{figure}
-
-In this example, the expressions \mintcode{(foo value1)} and \mintcode{(foo value2)} are both evaluated as \code{ForkJoinTask}s in a \code{ForkJoinPool}\cite{Lea2000}.
-The calls to \code{foo} are both forked immediately, then we attempt to evaluate the body of the \parlet{}.
-This means that the code in the body of the \code{let} which does not depend on the computations of \code{a} and \code{b} can execute without delay.
-Additionally, since the \code{ForkJoinPool} is designed for recursive workloads, tasks which are currently executing can create new tasks, submit them to the pool, then wait for the task to complete, without blocking execution.
-This means that nested \parlet{} forms (Figure \ref{fig:parlet_noblock}) will not block each other.
-
-\begin{figure}[h]
-\centering
-\RecustomVerbatimEnvironment{Verbatim}{BVerbatim}{}
-\begin{minted}{clojure}
-(parlet
-  [a (foo 100)]
-  ;; some other code not using a
-  (parlet
-    [b (foo 200)]
-    (+ a b)))
-\end{minted}
-\caption{Nested parlet forms}
-\label{fig:parlet_noblock}
-\end{figure}
-
-Code like this may not arise when the code is human generated, but it may arise when the code is generated by another macro.
-We will see some examples of this later.
-
-\subsection{Dependencies}
-The \parlet{} macro also supports simple dependency detection.
-Clojure \code{let} forms can use names defined previously in the let, and the bindings are evaluated from first to last.
-
-\begin{figure}[h]
-\centering
-\RecustomVerbatimEnvironment{Verbatim}{BVerbatim}{}
-\begin{minted}{clojure}
-(parlet
-  [a 1
+(parlet [a 1
          b (+ 1 a)]
@@
 If we plan on evaluating each binding in parallel, we can't allow the bindings to have dependencies.
-So, the \parlet{} macro looks through the bindings in the macro to determine if any of them depend on any of the names defined previously in the macro.
-If there are any, the macro will halt the compiler and report an error to the user.
+So, the \parlet{} macro will halt the compiler and report an error to the user if any dependencies are found in the \parlet{} form.
 
@@
 This simple dependency check, as well as the programmers promise that all function called are mostly pure Clojure code, allow us to ensure correct parallelism with this macro.
-
-% \section{parexpr}
-
-% The \parexpr{} macro breaks down expressions and (aggressively) evaluates them in parallel.
-% Suppose again that I had a long running function \code{foo} which I wanted to call twice, and add the results.
-% The code in Figure \ref{fig:parexpr} will make both of the long running calls to \code{foo} in parallel.
-% The \parexpr{} macro crawls the expression and expands it into multiple nested \parlet{} forms, filling each \parlet{} with as many evaluations as it can.
-
-% \begin{figure}[h]
-% \centering
-% \RecustomVerbatimEnvironment{Verbatim}{BVerbatim}{}
-% \begin{minted}{clojure}
-% (parexp (+ (foo value1) (foo value2)))
-% \end{minted}
-% \caption{Example using parexpr to evaluate long running functions}
-% \label{fig:parexpr}
-% \end{figure}
 
@@
 
-For an example see Figure \ref{fig:fib_code}.
+For an example see Figure \ref{fig:fib_parfun}.
 This defines a parallel Fibonacci function, which will only execute in parallel when the value of it's argument is greater than $35$.
 
-\begin{figure}[h]
+\begin{figure*}[t!]
+\begin{subfigure}[b]{0.5\textwidth}
+\RecustomVerbatimEnvironment{Verbatim}{BVerbatim}{}
+\centering
+\begin{minted}{clojure}
+(defn fib [n]
+  (if (or (= 0 n) (= 1 n))
+    1
+    (+
+     (fib (- n 1))
+     (fib (- n 2)))))
+\end{minted}
+\subcaption{Serial Fibonacci function}
+\label{fig:fib_serial}
+\end{subfigure}%
+\begin{subfigure}[b]{0.5\textwidth}
 \RecustomVerbatimEnvironment{Verbatim}{BVerbatim}{}
@@
 \end{minted}
-\caption{Fibonacci function with \defparfun{} added}
-\label{fig:fib_code}
-\end{figure}
+\subcaption{Fibonacci function with \defparfun{} added}
+\label{fig:fib_parfun}
+\end{subfigure}
+\caption{}
+\label{ref:fib_code}
+\end{figure*}
 
@@
 
-\begin{figure}[h]
-\RecustomVerbatimEnvironment{Verbatim}{BVerbatim}{}
-\centering
-\begin{minted}{clojure}
-(defun fib [n]
-  (if (or (= 0 n) (= 1 n))
-    1
-    (+
-     (fib (- n 1))
-     (fib (- n 2)))))
-\end{minted}
-\caption{Serial Fibonacci function}
-\label{fig:fib_serial}
-\end{figure}
-
 First we will look at the classical recursive Fibonacci example.
-Figure \ref{fig:fib_serial} shows the serial benchmark code; \ref{fig:fib_code} shows the code parallel benchmark code.
-In Figure \ref{fig:fibparfun} the results from many trials of this code running $1$, $2$, $4$, and $6$ cores.
+Figure \ref{fig:fib_serial} shows the serial benchmark code; \ref{fig:fib_parfun} shows the code parallel benchmark code.
+In Figure \ref{fig:cljfib} the results from many trials of this code running $1$, $2$, $4$, and $6$ cores.
 Each benchmark computes \mintcode{(fib 39)}.
@@
 
-<<fibparfun, fig.pos="H", fig.cap="Fibonacci Clojure (defparfun) Performance", echo=F, out.width='\\linewidth', fig.align='center'>>=
+<<cljfib, fig.pos="h", fig.cap="Fibonacci Clojure (defparfun) Performance", echo=F, out.width='\\linewidth', fig.align='center'>>=
 cljfib <- plot_means_error_for_benchmark(data, "fib", "parfun")
@@
 
-<<javafib, fig.pos="H", fig.cap="Fibonacci Java Performance", echo=F, out.width='\\linewidth', fig.align='center'>>=
+<<javafib, fig.pos="h", fig.cap="Fibonacci Java Performance", echo=F, out.width='\\linewidth', fig.align='center'>>=
 jfib <- plot_means_error_for_benchmark(data, "fib", "fj")
@@
 This effectively moves the stack for the recursive function to the heap.
-There is so upside to this; it eliminates the stack depth limit (which is an issue in Clojure because Clojure cannot implement tail call optimization), the large number of allocations creates a large amount garbage collector pressure.
+There is so upside to this; it eliminates the stack depth limit (which is an issue in Clojure because Clojure cannot implement tail call optimization), but the large number of allocations creates a large amount garbage collector pressure.
 The garbage collector behaves somewhat non---deterministically, so I believe this is the explanation for the large variation in runtime for the serial and recursive Fibonacci code.
@@
 
-<<fib_sd, fig.pos="H", fig.cap=" ", echo=F, out.width='\\linewidth', fig.align='center'>>=
+<<fib_sd, fig.pos="h", fig.cap="Fibonacci standard deviations", echo=F, out.width='\\linewidth', fig.align='center'>>=
 fib <- data[data$spec.name %like% "fib.big",]
@@
 barplot(t(as.matrix(d[,nnames])), beside=TRUE, names=d$cores, legend=nnames,
-        main = "Fibonacci standard deviations", xlab="cores", ylab="mean runtime standard deviation")
+        xlab="cores", ylab="mean runtime standard deviation") #, main = "Fibonacci standard deviations")
 @
@@
 
-<<id3parfun, fig.pos="H", fig.cap="id3 Clojure (defparfun) Performance", echo=F, out.width='\\linewidth', fig.align='center'>>=
+<<id3parfun, fig.pos="h", fig.cap="id3 Clojure (defparfun) Performance", echo=F, out.width='\\linewidth', fig.align='center'>>=
 id3 <- plot_means_error_for_benchmark(data, "id3", "parfun")
@@
 
+The intended users of these macros are developers whose primary concern is not performance, but may benefit from a simple mechanism with which they can take advantage of the many cores in their machines.
+Developers who are extremely concerned with performance and want a high degree of control should turn elsewhere (perhaps even to Java) to write their highly tuned code.
+
 \bibliography{/home/dpzmick/Documents/Bibtex/senior-thesis.bib}

### 2016-04-21 paper/paper.Rnw
updated paper

@@
 %\usemintedstyle{emacs}
-\usemintedstyle{borland}
+%\usemintedstyle{borland}
+\usemintedstyle{friendly}
 
 \usepackage{subcaption}
+\usepackage{caption}
 
@@
 As a result, there are a variety of libraries designed to allow developers to take advantage of the parallelism potential in Clojure.
-Clojure built ins such as \code{pmap}\footnotemark[1] and reducers\footnotemark[2] library provide data parallel sequence manipulation and transformation functions.
+Clojure built ins such as \code{pmap}\footnotemark[1] and \code{reducers}\footnotemark[2] library provide data parallel sequence manipulation and transformation functions.
 Third---party libraries like Tesser\footnotemark[3] and Claypoole\footnotemark[4] provide more data parallel APIs with slightly different goals than the builtin functions.
@@
 The \code{ForkJoinPool} library allows Java programmers to create lightweight, recursive tasks which are executed by a thread pool.
-Each thread in the pool maintains a queue of tasks to work on.
-Whenever a thread runs out of tasks to work on, it steals tasks from other threads.
+Each thread maintains a queue of tasks to work on.
+When a thread runs out of tasks, it steals tasks from other threads.
 Work stealing has been used a scheduling mechanism is a variety of modern threading libraries and has been very successful \cite{Lea2000, Blumofe1994, Blumofe1995a}
@@
 Clojure also has built in support for task parallelism via \code{future}\footnote{\url{https://clojuredocs.org/clojure.core/future}}.
-Each call to \code{future} spawns a thread to compute the result of a function call in the background.
-When a user is ready to access the computed value, they can \code{deref} the future to it.
+Each call to \code{future} spawns a thread to result of some function.
+When a user is ready to access the computed value, they \code{deref} the future.
 \code{deref} will block until the value is computed.
-For a user, \code{future} initially seem like a fairly natural way to parallelize recursive functions.
-Unfortunately, \code{future} only works well when a small number of tasks are created, or when the tasks do a lot of blocking I/O.
+For a user, \code{future} seem like a natural way to parallelize recursive functions.
+Unfortunately, \code{future} does not work well when a large number of tasks are created.
 Because thread creation overhead is high, users must be careful not to create an excessive number of threads or create threads which do too little work.
+
 Claypoole's implementation of \code{future} offers many improvements to the built in \code{future}.
-Their future implementation executes tasks created with \code{future} in a fixed sized thread pool, but, the \code{future} task will block the thread executing it if the task creates a new task, then immediately \code{deref}s it.
-This means that a programmer must be very careful not to create tasks in a Claypoole thread pool which depend on other tasks, otherwise their code may deadlock.
+The Claypoole implementation executes tasks in a fixed sized thread pool, but, the executing thread will block if the task creates a new task, then immediately \code{deref}s it.
+This means that a programmer must be very careful not to create tasks in a Claypoole thread pool which depend on other tasks that will also be submitted to the pool, otherwise their code may deadlock.
+
 The work in this paper is conceptually similar, but the interface I have used differs dramatically from the \code{future}s interface.
@@
 In Common Lisp, the \code{lparallel}\footnote{\url{https://lparallel.org/overview/}} library and macros are available.
-The macros I have implemented are very similar to the macros \code{lparallel} provides, but, Common Lisp programming conventions are not as ideal for these kinds of macros.
+The macros I have implemented are very similar to the macros \code{lparallel} provides, but, Common Lisp programming conventions are not as ideal for these kinds of macros, and Common Lisp programmers can't use the battle tested \code{ForkJoinPool} implementation.
 
@@
 In some cases, the order of certain side effects may not matter to a programmer.
-For example, it may not matter if we reorder \mintcode{(download file1)} and \mintcode{(download file2)} for a programmer writing a web scraper, but it may matter to a programmer writing a I/O constrained server.
+For example, a programmer writing a web scraper may not care what order \mintcode{(download file1)} and \mintcode{(download file2)} execute in, but the order may matter to a programmer writing a I/O constrained server.
 When a programmer feels that it may be acceptable to change the order of, or interleave, calls to mostly pure functions, we can reorder them subject to these constraints:
@@
 
-\tikzstyle{goodnode} = [rectangle, draw=blue, text width=0.8\linewidth]
+\tikzstyle{goodnode} = [rectangle, draw=black, text width=0.8\linewidth]
 \tikzstyle{badnode}  = [rectangle, draw=none,  text width=0.8\linewidth]
@@
     \node[coordinate] (0) at (1.0,0) {};
-    \node (start)       [badnode]                                 {\mintcode{(do}};
-    \node (a)           [badnode, below of=start, right of=start] {\mintcode{(let [a (bar 10)]}};
+    \node (a)           [badnode]                          {\mintcode{(let [a (bar 10)]}};
     \node (b)           [goodnode, below of=a, right of=a] {\mintcode{;; code that uses a}};
@@
 The first of the parallel macros is called \parlet{}.
-\parlet{} is a modified version of the Clojure \code{let} form.
 The \parlet{} macro has exactly the same behavior as Clojure's \code{let}, but it evaluates all of its bindings in parallel.
@@
 Additionally, since the \code{ForkJoinPool} is designed for recursive workloads, tasks which are currently executing can create new tasks, submit them to the pool, then wait for the task to complete, without blocking tasks created by other \parlet{} calls.
-So, \parlet{} forms can be nested; a programmer does not have to worry if functions called in the bindings of a \parlet{} form also use \parlet{}.
+This means that a programmer does not have to worry if functions called in the bindings of a \parlet{} form also use \parlet{}.
 
@@
 \section{defparfun}
+\begin{figure*}[t!]
+\captionsetup[subfigure]{belowskip=20pt}
+\begin{subfigure}[b]{0.5\textwidth}
+\RecustomVerbatimEnvironment{Verbatim}{BVerbatim}{}
+\centering
+\begin{minted}{clojure}
+(defn f [a]
+  (if (pred? ....)
+    (do ...)
+    (do ... (f (+ 1 a))))))
+\end{minted}
+\subcaption{}
+\label{fig:moveit_a}
+\end{subfigure}%
+\begin{subfigure}[b]{0.5\textwidth}
+\RecustomVerbatimEnvironment{Verbatim}{BVerbatim}{}
+\centering
+\begin{minted}{clojure}
+(defn f [a]
+  (if (pred? ....)
+    (do ...)
+    (let [e1 (f (+ 1 a))]
+      (do ... e1))))
+\end{minted}
+\subcaption{}
+\label{fig:moveit_b}
+\end{subfigure}
+\captionsetup[subfigure]{belowskip=2pt}
+\begin{subfigure}[b]{0.5\textwidth}
+\RecustomVerbatimEnvironment{Verbatim}{BVerbatim}{}
+\centering
+\begin{minted}{clojure}
+(defn f [...]
+  (let [a ...]
+    (let [b ...]
+      (f (+ 1 a)))))
+\end{minted}
+\subcaption{}
+\label{fig:moveit_c}
+\end{subfigure}%
+\begin{subfigure}[b]{0.5\textwidth}
+\RecustomVerbatimEnvironment{Verbatim}{BVerbatim}{}
+\centering
+\begin{minted}{clojure}
+(defn f [...]
+  (let [a ...]
+    (let [e1 (f (+ 1 a))]
+      (let [b ...] e1)))))
+\end{minted}
+\subcaption{}
+\label{fig:moveit_d}
+\end{subfigure}
+\caption{Moving a recursive function call}
+\label{fig:moveit}
+\end{figure*}
 
@@
 The \defparfun{} macro supports a granularity argument, allowing the programmer to specify when they would like to stop creating additional parallel tasks.
-The macro emits the expression provided for granularity inside of an \code{if} statement at the top of the function, so the programmer can use any arbitrary condition, including conditions dependent on the functions arguments, to decide when to stop spawning new tasks.
+The macro emits the expression provided for granularity inside of an \code{if} statement at the top of the function, so the programmer can use any arbitrary condition, including a condition dependent on the function's arguments, to decide when to stop spawning new tasks.
+Figure \ref{fig:fib_parfun} defines a parallel Fibonacci function, which will only execute in parallel when the value of it's argument is greater than $35$.
 
-For an example see Figure \ref{fig:fib_parfun}.
-This defines a parallel Fibonacci function, which will only execute in parallel when the value of it's argument is greater than $35$.
+\subsection{Implementation} \label{label:mpure_algo}
+In Clojure, any form which introduces control flow will eventually expand to an \code{if} for.
+Any form which introduces new bindings (including Clojure's destructuring mechanisms) will eventually expand to a \code{let} form.
+Because the constraints on movement of mostly pure functions only depend on control flow and variable bindings, we only need to make decisions about mostly pure function calls near \code{if} forms and \code{let} forms, in the full expanded versions of Clojure functions.
+
+When provided a function to manipulate, this function first expands all of the other macros in the function body, to get the nice property described above.
+Then, the macro recursively crawls the function body, looking for recursive call sites.
+If a call site is found, the recursive call (in the function body) is replaced by a newly introduced variable, and we hold onto the original recursive call.
+When all of the subexpressions for a given expression have been evaluated and transformed, we check if the expression is a \code{if} form or a \code{let} form.
+If we are sitting on a \code{if} form, all of the bindings introduced by the \code{true} branch and by the \code{false} branch are emitted in a new \code{let} expression (Figures \ref{fig:moveit_a} and \ref{fig:moveit_b}).
+This guarantees that condition 1 holds (Section \ref{label:mostly_pure}).
+If we are sitting on a \code{let} form, any of the bindings which depend on the expressions introduced by the \code{let} form are emitted.
+The remaining bindings continue to trickle upwards (see Figures \ref{fig:moveit_c} and \ref{fig:moveit_d}).
+This guarantees that condition 2 holds.
+
+After this transformation, it is possible to replace the \code{let} forms which bind the function results to their values with \parlet{} forms providing the same bindings.
+The introduction of the \parlet{} form introduces parallelism, so each recursive call will execute in the \code{ForkJoin} pool, in parallel.
+To complete the Fibonacci example, the \defparfun{} expansion is shown in Figure \ref{fig:parfun_result}.
+
+Because the transformation does not violate any of the properties defined for mostly pure functions, this transformation is safe when the function being declared as a \code{parfun} is mostly pure.
+
+\begin{figure}[h]
+\RecustomVerbatimEnvironment{Verbatim}{BVerbatim}{}
+\centering
+\begin{minted}{clojure}
+(defn fib [n]
+ ;; granularity check
+ (if (< n 35)
+  (if (or (= 0 n) (= 1 n))
+   1
+   (+ (fib (- n 1))
+      (fib (- n 2))))
+
+  ;; recursive case
+  (if (or (= 0 n) (= 1 n))
+   1
+   (parlet [expr17300 (fib (- n 1))
+            expr17301 (fib (- n 2))]
+    (+ expr17300 expr17301)))))
+\end{minted}
+\caption{Transformed Fibonacci function}
+\label{fig:parfun_result}
+\end{figure}
+
+\section{Benchmarking}
 
@@
 
-\subsection{Implementation} \label{label:mpure_algo}
-To do implement this macro, I first expand all of the \code{let} expressions in the function so that every \code{let} only has a single binding.
-Then, the syntax tree is crawled, finding every recursive function call.
-Every recursive call is replaced by a newly introduced variable.
-At the termination of each recursive call, the bindings introduced in the subexpression are returned.
-Once all of the subexpressions are evaluated, the algorithms decides if it must emit a \code{let} form for any of the bindings the subexpressions introduced.
-If the expression in question is an \code{if} statement, a let statement is introduced for the \code{true} branch and on the \code{false} branch, each containing all of the bindings introduced by the subexpressions for each branch.
-If the statement is a let statement, only the bindings which depend on the binding introduced by the let statement are emitted (following the let of course).
-This operation effectively pushed the evaluation of the function all as far ``up'' in the function as it can, following the constraints described in Section \ref{label:mostly_pure}.
-
-After this transformation, it is possible to replace the \code{let} forms which bind the function results to their values with \parlet{} forms providing the same bindings.
-The introduction of the \parlet{} form introduces parallelism, so each recursive call will execute in the \code{ForkJoin} pool, in parallel.
-
-Because the transformation does not violate any of the properties defined for mostly pure functions, this transformation is safe when the function being declared as a \code{parfun} is mostly pure.
-
-\begin{figure}[h]
-\RecustomVerbatimEnvironment{Verbatim}{BVerbatim}{}
-\centering
-\begin{minted}{clojure}
-(defn fib [n]
- ;; granularity check
- (if (< n 35)
-  (if (or (= 0 n) (= 1 n))
-   1
-   (+ (fib (- n 1))
-      (fib (- n 2))))
-
-  ;; recursive case
-  (if (or (= 0 n) (= 1 n))
-   1
-   (parlet [expr17300 (fib (- n 1))
-            expr17301 (fib (- n 2))]
-    (+ expr17300 expr17301)))))
-\end{minted}
-\caption{Transformed function}
-\label{fig:parfun_result}
-\end{figure}
-
-\section{Benchmarking}
 To run benchmarks, I used Google's Cloud Compute virtual machines.
@@
 
-To workaround the difficulties JVM benchmarking introduces, the Criterium\footnote{\url{https://github.com/hugoduncan/criterium}} library was used for Clojure code and the Caliper\footnote{\url{https://github.com/google/caliper}} library was used with Java code.
-The JVM does not expose a mechanism to control the total number of threads it creates (including garbage collection threads).
-Because each benchmark was run on it's own virtual machine with a constrained number of cores, the number of threads the JVM could create was controlled.
+To workaround the difficulties JVM benchmarking introduces, the Criterium\footnote{\url{https://github.com/hugoduncan/criterium}} library was used for Clojure code and the Caliper\footnote{\url{https://github.com/google/caliper}} library was used for Java code.
+Because each benchmark was run on it's own virtual machine with a constrained number of cores, the number of threads the JVM could create was controlled (including threads used for garbage collection)
 
@@
 When running on the \code{ForkJoinPool}, each function is further wrapped in a \code{RecursiveTask} object, causing additional allocations.
-This effectively moves the stack for the recursive function to the heap.
-There is so upside to this; it eliminates the stack depth limit (which is an issue in Clojure because Clojure cannot implement tail call optimization), but the large number of allocations creates a large amount garbage collector pressure.
+This effectively moves the stack for the recursive function to the heap (eliminating Clojure's stack depth limit\footnote{\url{http://clojure.org/about/functional\_programming\#\_recursive\_looping}}).
 The garbage collector behaves somewhat non---deterministically, so I believe this is the explanation for the large variation in runtime for the serial and recursive Fibonacci code.
@@
 I also implemented a simple ID3\footnote{\url{https://en.wikipedia.org/wiki/ID3_algorithm}} classifier in Clojure.
-The code is bit longer, so it is not included in this paper, but it can be found on this project's Github page\footnote{\url{https://github.com/dpzmick/auto_parallel}}.
-
+The code is bit longer, so it is not included in this paper, but it can be found on this project's GitHub page\footnote{\url{https://github.com/dpzmick/auto_parallel}}.
 For each benchmark, a random 1,000 element dataset was created.
@@
 
-The \code{ID3} code does much more work in each function call, so the additional overhead created by Clojure's function implementation does not seem to impact the results as much as it does in the Fibonacci benchmark.
-
+The \code{ID3} code does much more work in each function call, so the overhead introduced by the \code{ForkJoinPool} wrapper does not impact the results as much as it does in the Fibonacci benchmark.
 Figure \ref{fig:id3parfun} shows that we get the 3x speedup we expect on these virtual machines with the \code{ID3} algorithm.
@@
 
-It would also be useful to use profiling tools and implement static analysis tools which would detect potential locations for these macros.
-These tools could be used by developers to look for potential opportunities for parallelism, and they could be used to demonstrate the claim that many Clojure programs may benefit from use of these macros.
+% It would also be useful to use profiling tools and implement static analysis tools which would detect potential locations for these macros.
+% These tools could be used by developers to look for potential opportunities for parallelism, and they could be used to demonstrate the claim that many Clojure programs may benefit from use of these macros.
 
-The intended users of these macros are developers whose primary concern is not performance, but may benefit from a simple mechanism with which they can take advantage of the many cores in their machines.
-Developers who are extremely concerned with performance and want a high degree of control should turn elsewhere (perhaps even to Java) to write their highly tuned code.
+% The intended users of these macros are developers whose primary concern is not performance, but may benefit from a simple mechanism with which they can take advantage of the many cores in their machines.
+% Developers who are extremely concerned with performance and want a high degree of control should turn elsewhere (perhaps even to Java) to write their highly tuned code.
 

### 2016-04-22 paper/paper.Rnw
asd

@@
 When a thread runs out of tasks, it steals tasks from other threads.
-Work stealing has been used a scheduling mechanism is a variety of modern threading libraries and has been very successful \cite{Lea2000, Blumofe1994, Blumofe1995a}
+Work stealing has been used as a scheduling mechanism isn a variety of modern threading libraries and has been very successful \cite{Lea2000, Blumofe1994, Blumofe1995a}
 Users of a \code{ForkJoinPool} create a subclass of \code{RecursiveTask} to compute the value of some function.

### 2016-04-30 paper/paper.Rnw
updated paper

@@
 
+\newminted{clojure}{fontsize=\footnotesize}
+
 \author{
@@
 
-% question: Can macros be used to create interesting and useful parallelism
-% tools for recursive functions in clojure.
 \section{Introduction}
@@
 
-Using Clojure's macro system, I have implemented a set of macros which allow developers to take advantage of Clojure's task parallelism potential.
-I have shown that it is possible to attain reasonable degrees of parallelism with minimal modification to existing serial code.
+Using Clojure's macro system, we have implemented a set of macros which allow developers to take advantage of Clojure's task parallelism potential.
+We have shown that it is possible to attain reasonable degrees of parallelism with minimal modification to existing serial code.
 
@@
 Users of a \code{ForkJoinPool} create a subclass of \code{RecursiveTask} to compute the value of some function.
-Each of these tasks may create more \code{RecursiveTask}s, submit them to the \code{ForkJoinPool}, then wait for their subtasks to complete, without blocking the pool.
+Each of these tasks may create more \code{RecursiveTask}s, submit them to the \code{ForkJoinPool}, then wait for their subtasks to complete, without blocking any threads in the pool.
 Clojure programmers can use the \code{ForkJoinPool} libraries from Clojure, but the interface isn't exactly programmer friendly (or idiomatic).
@@
 The Claypoole implementation executes tasks in a fixed sized thread pool, but, the executing thread will block if the task creates a new task, then immediately \code{deref}s it.
-This means that a programmer must be very careful not to create tasks in a Claypoole thread pool which depend on other tasks that will also be submitted to the pool, otherwise their code may deadlock.
+This means that a programmer must be very careful not to create tasks in a Claypoole thread pool which depend on other tasks that will also be submitted to the pool.
 
-The work in this paper is conceptually similar, but the interface I have used differs dramatically from the \code{future}s interface.
+The work in this paper is conceptually similar, but the interface I have used differs dramatically from the \code{future} interface.
 My code also creates many tasks, but these tasks are created on a Java \code{ForkJoinPool}, so the tasks are much more lightweight, and they can be composed recursively.
@@
 In Common Lisp, the \code{lparallel}\footnote{\url{https://lparallel.org/overview/}} library and macros are available.
-The macros I have implemented are very similar to the macros \code{lparallel} provides, but, Common Lisp programming conventions are not as ideal for these kinds of macros, and Common Lisp programmers can't use the battle tested \code{ForkJoinPool} implementation.
+The macros I have implemented are very similar to the macros \code{lparallel} provides, but, Common Lisp programming conventions are not as ideal for these kinds of macros.
+Common Lisp programmers also do not have access to the battle tested \code{ForkJoinPool} implementation.
 
@@
 In other languages, frameworks like Cilk++\footnote{\url{https://www.cilkplus.org/}}, OpenMP\footnote{\url{http://openmp.org/wp/}}, Threading Building Blocks\footnote{\url{https://www.threadingbuildingblocks.org/}} have been implemented.
-Some of these require compiler support and advanced dependence analysis to guarantee correctness.
+Some of these require compiler support and complex dependence analysis to guarantee correctness.
 The macros I've implemented are an attempt to bring some of the niceties of these libraries and compiler extensions to Clojure, without the need to compiler support or complicated dependence analysis.
@@
 \section{Mostly Pure Functions} \label{label:mostly_pure}
-Before discussing the macros I've implemented, I need to loosely define a ``mostly pure'' function.
+Before discussing the macros we've implemented, we need to loosely define a ``mostly pure'' function.
 A mostly pure function is a thread---safe function with no side effects directly impacting the values of user-defined values, at the current level of abstraction.
-Mostly pure functions can be reordered (or interleaved) without impacting the values of user variables, although the change may impact I/O behavior and output order of a program.
-In some cases, the order of certain side effects may not matter to a programmer.
+A mostly pure function isn't actually pure (from a compiler's perspective), but, in many cases cases, the order of certain side effects may not matter to a programmer.
 For example, a programmer writing a web scraper may not care what order \mintcode{(download file1)} and \mintcode{(download file2)} execute in, but the order may matter to a programmer writing a I/O constrained server.
-When a programmer feels that it may be acceptable to change the order of, or interleave, calls to mostly pure functions, we can reorder them subject to these constraints:
+Mostly pure functions capture this idea; it may be the case that a function isn't actually pure, but we want to treat it as pure.
+Many Clojure functions fit this definition due to Clojure programming conventions and default immutable data structures.
+
+Calls to mostly pure functions can be reordered (or interleaved) without impacting the values of user variables, although the change may impact I/O behavior and output order of a program.
+When a programmer tells us a function is mostly pure, we can reorder them subject to the following constraints:
 
@@
 
-\tikzstyle{goodnode} = [rectangle, draw=black, text width=0.8\linewidth]
-\tikzstyle{badnode}  = [rectangle, draw=none,  text width=0.8\linewidth]
+\tikzstyle{goodround} = [ellipse, draw=black]
+\tikzstyle{badround}  = [ellipse, draw=black, style=dashed]
 
@@
 \centering
-\begin{tikzpicture}[node distance=0.62cm,auto]
-    \node[coordinate] (0) at (0.7,0) {};
-    \node (start0)      [badnode]                                        {\mintcode{;; cannot move the call outside}};
-    \node (start)       [badnode, below of=start0]                       {\mintcode{(if (pred? a b c ...)}};
-    \node (iftruedo)    [goodnode, below of=start, right of=start]       {\mintcode{(do}};
-    \node (iftruelet1)  [goodnode, below of=iftruedo, right of=iftruedo] {\mintcode{;; code that uses a, b, and/or c}};
-    \node (iftruelet2)  [goodnode, below of=iftruelet1]                  {\mintcode{(foo a b c))}};
-    \node (iffalse)     [badnode, below=2.8cm of 0]                      {\mintcode{(bar a b c))}};
+\begin{tikzpicture}[->, >=stealth', auto, node distance=0.8cm]
+    \tikzstyle{every node}=[font=\small]
+    \node (start) [badround] {\mintcode{(do ...}};
+    \node (startmiddle) [badround, below of=start] {\dots};
+    \node (startend) [badround, below of=startmiddle] {\mintcode{(if (pred? ...)}};
+
+    \draw[->] (start) -- (startmiddle);
+    \draw[->] (startmiddle) -- (startend);
+
+    \node (falsestart) [badround, below of=startend, left of=startend] {\mintcode{false}};
+    \node (falsemid) [badround, below of=falsestart] {\dots};
+    \node (falseend) [badround, below of=falsemid] {\mintcode{...}};
+
+    \draw[->] (startend) -- (falsestart);
+
+    \draw[->] (falsestart) -- (falsemid);
+    \draw[->] (falsemid) -- (falseend);
+
+    \node (truestart) [goodround, below of=startend, right of=startend] {\mintcode{true}};
+    \node (truemid) [goodround, below of=truestart] {\dots};
+    \node (trueend) [goodround, below of=truemid] {\mintcode{(f ...)}};
+
+    \draw[->] (startend) -- (truestart);
+
+    \draw[->] (truestart) -- (truemid);
+    \draw[->] (truemid) -- (trueend);
+
 \end{tikzpicture}
-\caption{The call to the mostly pure function \code{foo} can only be moved to the boxed nodes}
+\caption{The call to the mostly pure function \code{f} can only be moved to the nodes with solid borders}
 \label{fig:move_mostly_pure}
@@
 
+% \begin{figure}[h]
+% \centering
+% \begin{tikzpicture}[node distance=0.62cm,auto]
+%     \node[coordinate] (0) at (0.7,0) {};
+%     \node (start0)      [badnode]                                        {\mintcode{;; cannot move the call outside}};
+%     \node (start)       [badnode, below of=start0]                       {\mintcode{(if (pred? a b c ...)}};
+%     \node (iftruedo)    [goodnode, below of=start, right of=start]       {\mintcode{(do}};
+%     \node (iftruelet1)  [goodnode, below of=iftruedo, right of=iftruedo] {\mintcode{;; code that uses a, b, and/or c}};
+%     \node (iftruelet2)  [goodnode, below of=iftruelet1]                  {\mintcode{(foo a b c))}};
+%     \node (iffalse)     [badnode, below=2.8cm of 0]                      {\mintcode{(bar a b c))}};
+% \end{tikzpicture}
+% \caption{The call to the mostly pure function \code{foo} can only be moved to the boxed nodes}
+% \label{fig:move_mostly_pure}
+% \end{figure}
+
+% \begin{figure}[h]
+% \centering
+% \begin{tikzpicture}[node distance=0.61cm,auto]
+%     \node[coordinate] (0) at (1.0,0) {};
+%     \node (a)           [badnode]                          {\mintcode{(let [a (bar 10)]}};
+%     \node (b)           [goodnode, below of=a, right of=a] {\mintcode{;; code that uses a}};
+%     \node (c)           [goodnode, below of=b]             {\mintcode{(let [b (bar a)]}};
+%     \node (d)           [goodnode, below of=c, right of=c] {\mintcode{;; code that uses a and/or b}};
+%     \node (e)           [goodnode, below of=d]             {\mintcode{(foo a)))}};
+% \end{tikzpicture}
+% \caption{The call to the mostly pure function \code{foo} can only be moved to the boxed nodes}
+% \label{fig:move_mostly_pure_through_let}
+% \end{figure}
+
 \begin{figure}[h]
 \centering
-\begin{tikzpicture}[node distance=0.61cm,auto]
-    \node[coordinate] (0) at (1.0,0) {};
-    \node (a)           [badnode]                          {\mintcode{(let [a (bar 10)]}};
-    \node (b)           [goodnode, below of=a, right of=a] {\mintcode{;; code that uses a}};
-    \node (c)           [goodnode, below of=b]             {\mintcode{(let [b (bar a)]}};
-    \node (d)           [goodnode, below of=c, right of=c] {\mintcode{;; code that uses a and/or b}};
-    \node (e)           [goodnode, below of=d]             {\mintcode{(foo a)))}};
+\begin{tikzpicture}[->, >=stealth', auto, node distance=0.9cm]
+    \tikzstyle{every node}=[font=\small]
+    \node (start) [badround] {\mintcode{(do ...}};
+    \node (letstart) [badround, below of=start] {\mintcode{(let [a (...)]}};
+    \node (junk1) [goodround, below of=letstart, right of=letstart] {...};
+
+    \node (let2start) [goodround, below of=junk1] {\mintcode{(let [b (...)]}};
+    \node (call) [goodround, below of=let2start, right of=let2start] {\mintcode{(f a)}};
+
+    \draw[->] (start) -- (letstart);
+    \draw[->] (letstart) -- (junk1);
+    \draw[->] (junk1) -- (let2start);
+    \draw[->] (let2start) -- (call);
+
 \end{tikzpicture}
-\caption{The call to the mostly pure function \code{foo} can only be moved to the boxed nodes}
+\caption{The call to the mostly pure function \code{f} can only be moved to the nodes with solid borders}
 \label{fig:move_mostly_pure_through_let}
 \end{figure}
-
-Many Clojure functions fit this definition due to Clojure programming conventions and default immutable data structures.
-In this paper, we rely on the judgment of the programmer to tell us when a function can be considered mostly pure.
 
@@
 \RecustomVerbatimEnvironment{Verbatim}{BVerbatim}{}
-\begin{minted}{clojure}
+\begin{minted}[fontsize=\small]{clojure}
 (parlet
@@
 \RecustomVerbatimEnvironment{Verbatim}{BVerbatim}{}
-\begin{minted}{clojure}
+\begin{minted}[fontsize=\small]{clojure}
 (let
@@
 \RecustomVerbatimEnvironment{Verbatim}{BVerbatim}{}
-\begin{minted}{clojure}
+\begin{minted}[fontsize=\small]{clojure}
 (parlet [a 1
@@
 If the programmer chooses to use a \parlet{} form, we assume that the functions called in the bindings are mostly pure.
-This simple dependency check, as well as the programmers promise that all function called are mostly pure Clojure code, allow us to ensure correct parallelism with this macro.
+The simple dependency check combined with the programmers promise that all function called are mostly pure allow us to ensure correct parallelism with this macro.
 
@@
 \captionsetup[subfigure]{belowskip=20pt}
-\begin{subfigure}[b]{0.5\textwidth}
+\begin{subfigure}[b]{0.33\textwidth}
 \RecustomVerbatimEnvironment{Verbatim}{BVerbatim}{}
 \centering
-\begin{minted}{clojure}
+\begin{minted}[fontsize=\small]{clojure}
 (defn f [a]
@@
     (do ...)
-    (do ... (f (+ 1 a))))))
+    (do ... (f (+ 1 a)))))
 \end{minted}
-\subcaption{}
+\subcaption{Input code}
 \label{fig:moveit_a}
 \end{subfigure}%
-\begin{subfigure}[b]{0.5\textwidth}
+\begin{subfigure}[b]{0.33\textwidth}
 \RecustomVerbatimEnvironment{Verbatim}{BVerbatim}{}
 \centering
-\begin{minted}{clojure}
+\begin{minted}[fontsize=\small]{clojure}
+(defn f [a]
+  (if (pred? ....)
+    (do ...)
+    (do ... e1)))
+\end{minted}
+\subcaption{Replace call with new variable}
+\label{fig:moveit_b}
+\end{subfigure}%
+\begin{subfigure}[b]{0.33\textwidth}
+\RecustomVerbatimEnvironment{Verbatim}{BVerbatim}{}
+\centering
+\begin{minted}[fontsize=\small]{clojure}
 (defn f [a]
@@
 \end{minted}
-\subcaption{}
-\label{fig:moveit_b}
+\subcaption{The \code{if} forces a \code{let}}
+\label{fig:moveit_c}
 \end{subfigure}
 \captionsetup[subfigure]{belowskip=2pt}
-\begin{subfigure}[b]{0.5\textwidth}
+\begin{subfigure}[b]{0.33\textwidth}
 \RecustomVerbatimEnvironment{Verbatim}{BVerbatim}{}
 \centering
-\begin{minted}{clojure}
+\begin{minted}[fontsize=\small]{clojure}
 (defn f [...]
@@
 \end{minted}
-\subcaption{}
-\label{fig:moveit_c}
+\subcaption{Input code}
+\label{fig:moveit_d}
 \end{subfigure}%
-\begin{subfigure}[b]{0.5\textwidth}
+\begin{subfigure}[b]{0.33\textwidth}
 \RecustomVerbatimEnvironment{Verbatim}{BVerbatim}{}
 \centering
-\begin{minted}{clojure}
+\begin{minted}[fontsize=\small]{clojure}
+(defn f [...]
+  (let [a ...]
+    (let [b ...]
+      e1)))
+\end{minted}
+\subcaption{Introduce a new variable}
+\label{fig:moveit_e}
+\end{subfigure}%
+\begin{subfigure}[b]{0.33\textwidth}
+\RecustomVerbatimEnvironment{Verbatim}{BVerbatim}{}
+\centering
+\begin{minted}[fontsize=\small]{clojure}
 (defn f [...]
@@
     (let [e1 (f (+ 1 a))]
-      (let [b ...] e1)))))
+      (let [b ...]
+        e1)))))
 \end{minted}
-\subcaption{}
-\label{fig:moveit_d}
+\subcaption{The \code{let} which defines \code{a} forces a \code{let}}
+\label{fig:moveit_f}
 \end{subfigure}
@@
 
-\defparfun{} allows a programmer to parallelize calls to a recursive function.
+\defparfun{} allows a programmer define a function which will parallelize its recursive calls.
 The \defparfun{} macro supports a granularity argument, allowing the programmer to specify when they would like to stop creating additional parallel tasks.
@@
 Figure \ref{fig:fib_parfun} defines a parallel Fibonacci function, which will only execute in parallel when the value of it's argument is greater than $35$.
+If a programmer choses to use \defparfun{} to define a function, we assume that the function being defined is mostly pure.
 
 \subsection{Implementation} \label{label:mpure_algo}
-In Clojure, any form which introduces control flow will eventually expand to an \code{if} for.
+In Clojure, any form which introduces control flow will eventually expand to an \code{if} form.
 Any form which introduces new bindings (including Clojure's destructuring mechanisms) will eventually expand to a \code{let} form.
-Because the constraints on movement of mostly pure functions only depend on control flow and variable bindings, we only need to make decisions about mostly pure function calls near \code{if} forms and \code{let} forms, in the full expanded versions of Clojure functions.
+Because the constraints on movement of mostly pure functions only depend on control flow and variable bindings, we only need to make decisions about mostly pure function calls near \code{if} forms and \code{let} forms, in the fully expanded code for a function.
 
-When provided a function to manipulate, this function first expands all of the other macros in the function body, to get the nice property described above.
+When provided a function to manipulate, this macro first expands all of the other macros in the function body, to get the nice property described above.
 Then, the macro recursively crawls the function body, looking for recursive call sites.
@@
 When all of the subexpressions for a given expression have been evaluated and transformed, we check if the expression is a \code{if} form or a \code{let} form.
-If we are sitting on a \code{if} form, all of the bindings introduced by the \code{true} branch and by the \code{false} branch are emitted in a new \code{let} expression (Figures \ref{fig:moveit_a} and \ref{fig:moveit_b}).
+If we are sitting on a \code{if} form, all of the bindings introduced by the \code{true} branch and by the \code{false} branch are emitted in a new \code{let} expression (Figures \ref{fig:moveit_a}, \ref{fig:moveit_b}, and \ref{fig:moveit_c}).
 This guarantees that condition 1 holds (Section \ref{label:mostly_pure}).
 If we are sitting on a \code{let} form, any of the bindings which depend on the expressions introduced by the \code{let} form are emitted.
-The remaining bindings continue to trickle upwards (see Figures \ref{fig:moveit_c} and \ref{fig:moveit_d}).
+The remaining bindings continue to trickle upwards (see Figures \ref{fig:moveit_d}, \ref{fig:moveit_e}, and \ref{fig:moveit_f}).
 This guarantees that condition 2 holds.
@@
 
-Because the transformation does not violate any of the properties defined for mostly pure functions, this transformation is safe when the function being declared as a \code{parfun} is mostly pure.
+Because the transformation does not violate any of the properties defined for mostly pure functions, this transformation is safe.
 
@@
 \centering
-\begin{minted}{clojure}
+\begin{minted}[fontsize=\small]{clojure}
 (defn fib [n]
@@
 \centering
-\begin{minted}{clojure}
+\begin{minted}[fontsize=\small]{clojure}
 (defparfun fib [n] (< n 35)
@@
 Since it does not appear in the Java benchmarks, it is not a result of variable performance in the Google Cloud Platform virtual machines.
-In Figure \ref{fig:fib_sd} I show the standard deviation of the mean runtime for the serial and parallel Clojure Fibonacci functions, along with their Java counterparts.
+Figure \ref{fig:fib_sd} shows the standard deviations of the mean runtime for the serial and parallel Clojure Fibonacci functions, along with their Java counterparts.
 Notice that the Java results do not have nearly as high deviations from mean runtime.
-While I cannot completely explain the variability, it seems to be caused by the increased pressure Clojure's function implementation and the \code{ForkJoinPool} wrapper tasks places on the JVM garbage collector.
+While we cannot completely explain the variability, it seems to be caused by the increased pressure Clojure's function implementation and the \code{ForkJoinPool} wrapper tasks places on the JVM garbage collector.
 Every Clojure function is an \code{Object} created following the Clojure \code{IFn} interface\footnote{\url{http://clojure.org/reference/special_forms\#fn}}.
@@
 This effectively moves the stack for the recursive function to the heap (eliminating Clojure's stack depth limit\footnote{\url{http://clojure.org/about/functional\_programming\#\_recursive\_looping}}).
-The garbage collector behaves somewhat non---deterministically, so I believe this is the explanation for the large variation in runtime for the serial and recursive Fibonacci code.
+The garbage collector behaves somewhat non---deterministically, so we believe that this is the explanation for the large variations in runtime present in the Clojure results.
 We can avoid excessive task creation by controlling the granularity of parallelism, to an extent.
@@
 \subsection{ID3} \label{label:id3_bench}
-I also implemented a simple ID3\footnote{\url{https://en.wikipedia.org/wiki/ID3_algorithm}} classifier in Clojure.
+We also implemented a simple ID3\footnote{\url{https://en.wikipedia.org/wiki/ID3_algorithm}} classifier in Clojure.
 The code is bit longer, so it is not included in this paper, but it can be found on this project's GitHub page\footnote{\url{https://github.com/dpzmick/auto_parallel}}.
@@
 \section{Conclusions and Future Work}
-The Clojure macros I've implemented perform transformation which can speedup Clojure code to a degree which matches the speedups attained using handwritten Java code, running on the same hardware.
+The Clojure macros we've implemented perform transformations which can speed up serial Clojure code to a degree which matches the speedups attained using handwritten Java code, running on the same hardware.
 Parallelism is difficult, and automatic parallelism is possible\cite{Banerjee1993}, but these techniques are complicated and often do not get the desired results and the research community has begun to feel the need for explicit parallelism in programs.\cite{Arvind2010}
-Languages like Clojure are well suited for this parallelism and techniques like mine can easily be implemented.
-In a language with a strong STM system and immutable structures, such transformations are easy to reason about, making it much simpler for programmers to implement explicitly parallel programs.
+These macros are designed to allow programmers whose primary concern is not performance to write explicitly parallel programs.
+In a language with a strong STM system and immutable structures, macros like these are easy to reason about, making it much simpler for programmers to implement explicitly parallel programs.
 
 Macros of this style do not inhibit the programmers ability to use the other mechanisms implemented in the language, although interoperability with them could be improved.
-For example, one of the tests which was not discussed in this paper involved using the STM system from within a function declared with \defparfun{}.
+For example, one of the tests which was not discussed in this paper used the STM system from within a function declared with \defparfun{}.
 Benchmarks on this code behaved correctly and performance improved as expected.
 However, if a programmer attempted to use a \code{pmap} or \code{future} inside of a \defparfun{} or \parlet{}, the two systems would create separate thread pools and the number of created threads would be large, possibly causing poor performance.
-There are also a variety of other useful macros in lparallel\footnote{\url{https://lparallel.org/}} that may be useful to implement in Clojure and would complement the macros I've implemented in this project.
+There are also a variety of other useful macros in lparallel\footnote{\url{https://lparallel.org/}} would be useful to implement in Clojure.
 
